---
title: "Arteries classtpx vs SVM"
author: "Kushal K Dey"
date: "April 29, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Overview

In this script, we extract tissue samples from two tissues in the GTEx V6 samples and apply a classification due to Support Vector Machines (SVM) and the classtpx model. We would like to figure out which of the two methods has higher misclassification rate.

```{r warning=FALSE, message=FALSE}
library(data.table)
library(e1071)
library(classtpx)
library(maptpx)
```

## Data Preparation (Arteries)

```{r message=FALSE, warning=FALSE}
gtex_data <- data.frame(fread("../data/GTEX_V6/cis_gene_expression.txt"));
matdata <- gtex_data[,-(1:2)];

tissue_labels=read.table("../data/GTEX_V6/samples_id.txt")[,3];

aorta_labels <- which(as.character(tissue_labels) == 'Artery - Aorta')
coronary_labels <- which(as.character(tissue_labels) == 'Artery - Coronary');
tibial_labels <- which(as.character(tissue_labels) == 'Artery - Tibial');

pooled_data <- t(cbind(matdata[,aorta_labels],
                       matdata[,coronary_labels],
                       matdata[,tibial_labels]))

```

## maptpx application

We first perform `maptpx`.

```{r echo=TRUE, eval=FALSE}
Topic_clus <- maptpx::topics(pooled_data, K=3, tol=100);
save(Topic_clus, file="../rdas/gtex_arteries.rda");
```

We plot suing the Structure plot visualization.

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/gtex_arteries.rda"));

par(mfrow=c(1,1))
par(mar=c(4,2,2,2))
omega <- Topic_clus$omega;
annotation <- data.frame(
sample_id = paste0("X", c(1:NROW(omega))),
tissue_label = factor(tissue_labels[c(aorta_labels, 
                                      coronary_labels,
                                      tibial_labels)],
                      levels=c("Artery - Aorta", 
                               "Artery - Coronary",
                               "Artery - Tibial"))
)

rownames(omega) <- annotation$sample_id;
CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))

```

## Support Vector Machine (SVM) application 

We filter out the first $50$ samples from the Blood samples and the first $50$ samples from the Pancreas tissues and then perform classification of the other samples based on these two sets of samples. 

### On counts

```{r echo=TRUE, eval=TRUE}

gene_names <- read.table("../data/GTEX_V6/gene_names_GTEX_V6.txt");
gene_names <- substring(gene_names[,1],1,15);

training_data <- t(cbind(matdata[,aorta_labels[1:50]],
                         matdata[,coronary_labels[1:50]],
                         matdata[,tibial_labels[1:50]]))

test_data <- t(cbind(matdata[,aorta_labels[-(1:50)]],
                         matdata[,coronary_labels[-(1:50)]],
                         matdata[,tibial_labels[-(1:50)]]))

tissue_type <- c(rep("Aorta",50), 
                 rep("Coronary",50),
                 rep("Tibial", 50));

training.data.frame <- cbind.data.frame(training_data, tissue_type);
colnames(training.data.frame) <- c(gene_names, "tissue_type");

model <- svm(tissue_type ~ ., data = training.data.frame)

test_class_svm <- predict(model, test_data)

tab_class_svm <- table(test_class_svm, 
                c(rep("Aorta", length(aorta_labels)-50),   
                  rep("Coronary", length(coronary_labels)-50),
                  rep("Tibial", length(tibial_labels)-50)))

print(tab_class_svm)

misclass_svm <- sum(tab_class_svm[row(tab_class_svm)!=col(tab_class_svm)])/ sum(tab_class_svm)
misclass_svm
```


### On voom counts

```{r echo=TRUE, eval=TRUE}

gene_names <- read.table("../data/GTEX_V6/gene_names_GTEX_V6.txt");
gene_names <- substring(gene_names[,1],1,15);

voom_matdata <- data.frame(fread("../data/GTEX_V6/voom_cis_gene_expression.txt"))
voom_matdata <- voom_matdata[,-1]

training_data <- t(cbind(voom_matdata[,aorta_labels[1:50]],
                         voom_matdata[,coronary_labels[1:50]],
                         voom_matdata[,tibial_labels[1:50]]))

test_data <- t(cbind(voom_matdata[,aorta_labels[-(1:50)]],
                         voom_matdata[,coronary_labels[-(1:50)]],
                         voom_matdata[,tibial_labels[-(1:50)]]))

tissue_type <- c(rep("Aorta",50), 
                 rep("Coronary",50),
                 rep("Tibial", 50));

training.data.frame <- cbind.data.frame(training_data, tissue_type);
colnames(training.data.frame) <- c(gene_names, "tissue_type");

model <- svm(tissue_type ~ ., data = training.data.frame)

test_class_svm <- predict(model, test_data)

tab_class_svm <- table(test_class_svm, 
                c(rep("Aorta", length(aorta_labels)-50),   
                  rep("Coronary", length(coronary_labels)-50),
                  rep("Tibial", length(tibial_labels)-50)))

print(tab_class_svm)

misclass_svm <- sum(tab_class_svm[row(tab_class_svm)!=col(tab_class_svm)])/ sum(tab_class_svm)
misclass_svm
```

## classtpx (theta.fix) application

```{r echo=TRUE, eval=TRUE}
class_labs <- c(rep(1,50), rep(2,50), rep(3,50));
known_samples <- c(1:50, 
                   length(aorta_labels) + (1:50), 
                   length(aorta_labels)+length(coronary_labels)+ (1:50));

```


```{r echo=TRUE, eval=FALSE}
Topic_clus <- classtpx::class_topics(
    pooled_data, 
    K=3, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    shrink=TRUE,
    shrink.method = 1,
    tol=0.0001,
    ord=FALSE)

save(Topic_clus, file="../rdas/arteries_theta_fix_classtpx_voom.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/arteries_theta_fix_classtpx_voom.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Aorta", "Coronary", "Tibial")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(
      rep("Aorta", length(aorta_labels)), 
      rep("Coronary", length(coronary_labels)),
      rep("Tibial", length(tibial_labels)))[-known_samples],
      test_class_classtpx[-known_samples]);

print(tab_class_classtpx)
misclass_classtpx <- sum(tab_class_classtpx[row(tab_class_classtpx)!=col(tab_class_classtpx)])/ sum(tab_class_classtpx)
misclass_classtpx

print(tab_class_classtpx)
```

```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,50), rep(2,50), rep(3,50));
known_samples <- c(1:50, 
                   length(aorta_labels) + (1:50), 
                   length(aorta_labels)+length(coronary_labels)+ (1:50));

Topic_clus <- class_topics(
    pooled_data, 
    K=3, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    shrink=TRUE,
    shrink.method = 2,
    tol=0.0001,
    ord=FALSE)

save(Topic_clus, file="../rdas/arteries_theta_fix_classtpx.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/arteries_theta_fix_classtpx.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Aorta", "Coronary", "Tibial")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(
      rep("Aorta", length(aorta_labels)), 
      rep("Coronary", length(coronary_labels)),
      rep("Tibial", length(tibial_labels)))[-known_samples],
      test_class_classtpx[-known_samples]);

print(tab_class_classtpx)
misclass_classtpx <- sum(tab_class_classtpx[row(tab_class_classtpx)!=col(tab_class_classtpx)])/ sum(tab_class_classtpx)
misclass_classtpx

print(tab_class_classtpx)
```


```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,50), rep(2,50), rep(3,50));
known_samples <- c(1:50, 
                   length(aorta_labels) + (1:50), 
                   length(aorta_labels)+length(coronary_labels)+ (1:50));

Topic_clus <- classtpx::class_topics(
    pooled_data, 
    K=3, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    shrink=FALSE,
    shrink.method = 1,
    tol=0.0001,
    ord=FALSE)

save(Topic_clus, file="../rdas/arteries_theta_fix_classtpx_no_shrink.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/arteries_theta_fix_classtpx_no_shrink.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Aorta", "Coronary", "Tibial")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(
      rep("Aorta", length(aorta_labels)), 
      rep("Coronary", length(coronary_labels)),
      rep("Tibial", length(tibial_labels)))[-known_samples],
      test_class_classtpx[-known_samples]);

print(tab_class_classtpx)
misclass_classtpx <- sum(tab_class_classtpx[row(tab_class_classtpx)!=col(tab_class_classtpx)])/ sum(tab_class_classtpx)
misclass_classtpx

print(tab_class_classtpx)
```

```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,50), rep(2,50), rep(3,50));
known_samples <- c(1:50, 
                   length(aorta_labels) + (1:50), 
                   length(aorta_labels)+length(coronary_labels)+ (1:50));

Topic_clus <- classtpx::class_topics(
    pooled_data, 
    K=3, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="omega.fix",
    shrink=FALSE,
    shrink.method = 2,
    tol=0.0001,
    ord=FALSE)

save(Topic_clus, file="../rdas/arteries_omega_fix_classtpx.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/arteries_omega_fix_classtpx.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Aorta", "Coronary", "Tibial")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(
      rep("Aorta", length(aorta_labels)), 
      rep("Coronary", length(coronary_labels)),
      rep("Tibial", length(tibial_labels)))[-known_samples],
      test_class_classtpx[-known_samples]);

print(tab_class_classtpx)
misclass_classtpx <- sum(tab_class_classtpx[row(tab_class_classtpx)!=col(tab_class_classtpx)])/ sum(tab_class_classtpx)
misclass_classtpx

print(tab_class_classtpx)
```
