---
title: 'Batch correction: Simulation check'
author: "Kushal K Dey & Joyce Hsiao"
date: "September 24, 2015"
output: 
  html_document:
    toc: true
---

```{r echo=FALSE, eval=TRUE}
rm(list=ls())
```

```{r echo=TRUE, eval=TRUE}
library(CountClust)
suppressMessages(library(mygene))
library(knitr)
suppressMessages(library(data.table))
suppressMessages(library(maptpx))
```


## Objective

In this script, we provide a simulation check for the Batch correction algorithm we are using. We perform two simulation studies to obtain two count tables and then apply topic model before and after batch correction. 

## Simulation model 1

```{r}

simulate_poisson_random_batch_1 = function(n_samples,
                                          n_genes, 
                                          n_clus,
                                          topic_omega_sim,
                                          topic_alpha_sim,
                                          topic_beta_sim,
                                          label_batch_sim)
{
  
  # n_samples : the number of tissue or single cell samples
  # n_genes : the number of genes
  # n_clus: the number of topics/clusters
  # topic_omega_sim: omega (n_samples \times n_clus ) matrix of mixing proportions
  # topic_alpha_sim: alpha (n_clus \times n_genes) matrix of topic expressions
  # topic_beta_sim : beta (num_batches \times n_genes) matrix of batch effects
  # label_batch_sim: the labels of the batches startinf grom 1 to num_batches
  
  flag <- 0;
  if(n_clus !=dim(topic_omega_sim)[2]){
    print ("The number of clusters does not match with dimension of omega matrix");
    flag <- 1}
  if(n_clus !=dim(topic_alpha_sim)[1]){
    print ("The number of clusters does not match with dimension of alpha matrix");
    flag <- 1}
  if(max(label_batch_sim) !=dim(topic_beta_sim)[1]){
    print ("The number of batches does not match with the label vector");
    flag <- 1}
  if(flag==0) {
    out_temp=exp(topic_omega_sim%*%topic_alpha_sim +topic_beta_sim[label_batch_sim,]);
    out_temp_vec <- as.vector(out_temp);
    read_counts=matrix(unlist(lapply(1:length(out_temp_vec),function(x) rpois(1,out_temp_vec[x]))),nrow=n_samples);
    return(read_counts)}
  
}


N_sim = 500;
N_genes =1000;
N_clus =4;

alpha_true=matrix(rnorm(N_clus*N_genes,0.5,1),nrow=N_clus); ### the matrix of fixed effects

lab_batch=c(rep(1,N_sim/2),rep(2,N_sim/2));

B= max(lab_batch);

sigmab_true=2;

beta_true=matrix(0,B,N_genes);       ###  the matrix of the random effect

for(g in 1:N_genes)
{
  beta_true[,g]=rnorm(B,mean=0,sd=sigmab_true);
}

library(gtools)
T=N_sim/5;
omega_true=matrix(rbind(rdirichlet(T,c(3,4,2,6)),
                        rdirichlet(T,c(1,4,6,3)),
                        rdirichlet(T,c(4,1,2,2)),
                        rdirichlet(T,c(2,6,3,2)),
                        rdirichlet(T,c(3,3,5,4))), nrow=N_sim);


counts <- simulate_poisson_random_batch_1(N_sim, N_genes, N_clus, omega_true, alpha_true, beta_true, lab_batch)


```


###  True structure plot

```{r echo=TRUE, eval=TRUE}

barplot(t(omega_true),col=2:(N_clus+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",N_clus),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```


### Topic model fit (batch uncorrected)

```{r echo=TRUE, eval=TRUE, cache=TRUE}

Topic_clus <- topics(counts,K=4, tol=0.5);
docweights_topics=Topic_clus$omega;
barplot(t(docweights_topics),col=2:(N_clus+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",N_clus),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

### Batch corrected Topic model fit 

```{r echo=TRUE, eval=TRUE, cache=TRUE}

batch_corrected_counts <- BatchCorrectedCounts(counts, lab_batch);

Topic_clus <- topics(batch_corrected_counts,K=4, tol=0.5);
docweights_topics=Topic_clus$omega;

library(permute);
library("BioPhysConnectoR");
perm_set=rbind(1:N_clus,allPerms(1:N_clus));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
  temp=docweights_topics[,perm_set[p,]];
  diff[p]=fnorm(temp,omega_true);
}

p_star=which(diff==min(diff));
docweights_topics=docweights_topics[,perm_set[p_star,]];


barplot(t(docweights_topics),col=2:(N_clus+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",N_clus),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

We observe that the topic model fit on the batch corrected counts is pretty close to the original Structure plot. This sort of validates the fact that the batch correction scheme is working.

## Simulation model 2

```{r}

simulate_poisson_random_batch_2 = function(n_samples,
                                          n_genes, 
                                          n_clus,
                                          topic_omega_sim,
                                          topic_alpha_sim,
                                          topic_beta_sim,
                                          label_batch_sim)
{
  
  # n_samples : the number of tissue or single cell samples
  # n_genes : the number of genes
  # n_clus: the number of topics/clusters
  # topic_omega_sim: omega (n_samples \times n_clus ) matrix of mixing proportions
  # topic_alpha_sim: alpha (n_clus \times n_genes) matrix of topic expressions
  # topic_beta_sim : beta (num_batches \times n_genes) matrix of batch effects
  # label_batch_sim: the labels of the batches startinf grom 1 to num_batches
  
  flag <- 0;
  if(n_clus !=dim(topic_omega_sim)[2]){
    print ("The number of clusters does not match with dimension of omega matrix");
    flag <- 1}
  if(n_clus !=dim(topic_alpha_sim)[1]){
    print ("The number of clusters does not match with dimension of alpha matrix");
    flag <- 1}
  if(max(label_batch_sim) !=dim(topic_beta_sim)[1]){
    print ("The number of batches does not match with the label vector");
    flag <- 1}
  if(flag==0) {
    out_temp <- matrix(0,nrow=n_samples, ncol=n_genes);
    for(n in 1:n_samples)
    {
      out_temp[n,] <- topic_omega_sim[n,]%*% exp(topic_alpha_sim + topic_beta_sim[label_batch_sim[n],]);
    }
    out_temp_vec <- as.vector(out_temp);
    read_counts=matrix(unlist(lapply(1:length(out_temp_vec),function(x) rpois(1,out_temp_vec[x]))),nrow=n_samples);
    return(read_counts)}
  
}


N_sim = 500;
N_genes =1000;
N_clus =4;

alpha_true=matrix(rnorm(N_clus*N_genes,0.5,1),nrow=N_clus); ### the matrix of fixed effects

lab_batch=c(rep(1,N_sim/2),rep(2,N_sim/2));

B= max(lab_batch);

sigmab_true=2;

beta_true=matrix(0,B,N_genes);       ###  the matrix of the random effect

for(g in 1:N_genes)
{
  beta_true[,g]=rnorm(B,mean=0,sd=sigmab_true);
}

library(gtools)
T=N_sim/5;
omega_true=matrix(rbind(rdirichlet(T,c(3,4,2,6)),
                        rdirichlet(T,c(1,4,6,3)),
                        rdirichlet(T,c(4,1,2,2)),
                        rdirichlet(T,c(2,6,3,2)),
                        rdirichlet(T,c(3,3,5,4))), nrow=N_sim);


counts <- simulate_poisson_random_batch_2(N_sim, N_genes, N_clus, omega_true, alpha_true, beta_true, lab_batch)


```

###  True structure plot

```{r echo=TRUE, eval=TRUE}

barplot(t(omega_true),col=2:(N_clus+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",N_clus),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

### Topic model fit (batch uncorrected)

```{r echo=TRUE, eval=TRUE, cache=TRUE}

Topic_clus <- topics(counts,K=4, tol=0.5);
docweights_topics=Topic_clus$omega;
barplot(t(docweights_topics),col=2:(N_clus+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",N_clus),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

### Batch corrected Topic model fit 

```{r echo=TRUE, eval=TRUE, cache=TRUE}

batch_corrected_counts <- BatchCorrectedCounts(counts, lab_batch);

Topic_clus <- topics(batch_corrected_counts,K=4, tol=0.5);
docweights_topics=Topic_clus$omega;

library(permute);
library("BioPhysConnectoR");
perm_set=rbind(1:N_clus,allPerms(1:N_clus));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
  temp=docweights_topics[,perm_set[p,]];
  diff[p]=fnorm(temp,omega_true);
}

p_star=which(diff==min(diff));
docweights_topics=docweights_topics[,perm_set[p_star,]];


barplot(t(docweights_topics),col=2:(N_clus+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",N_clus),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```


```{r }
sessionInfo()
```