---
title: 'classtpx vs SVM on GTEx V6 data: Brain cerebellum vs cerebellar hemisphere'
author: "Kushal K Dey"
date: "April 7, 2016"
output: html_document
---

## Overview

In this script, we extract tissue samples from two tissues in the GTEx V6 samples and apply a classification due to Support Vector Machines (SVM) and the classtpx model. We would like to figure out which of the two methods has higher misclassification rate.

```{r warning=FALSE, message=FALSE}
library(data.table)
library(e1071)
library(classtpx)
library(maptpx)
```

## Data preparation (Cerebellum + Cerebellar Hemisphere)

```{r message=FALSE, warning=FALSE}
gtex_data <- data.frame(fread("../data/GTEX_V6/cis_gene_expression.txt"));
matdata <- gtex_data[,-(1:2)];

tissue_labels=read.table("../data/GTEX_V6/samples_id.txt")[,3];

cerebellum_labels <- which(as.character(tissue_labels) == 'Brain - Cerebellum')
hemisphere_labels <- which(as.character(tissue_labels) == 'Brain - Cerebellar Hemisphere');

pooled_data <- t(cbind(matdata[,cerebellum_labels],matdata[,hemisphere_labels]))
```

## maptpx application

We first perform `maptpx`.

```{r echo=TRUE, eval=FALSE}
Topic_clus <- maptpx::topics(pooled_data, K=2, tol=0.1);
save(Topic_clus, file="../rdas/gtex_cerebellum_hemisphere.rda");
```

We plot suing the Structure plot visualization.

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/gtex_cerebellum_hemisphere.rda"));

par(mfrow=c(1,1))
par(mar=c(4,2,2,2))
omega <- Topic_clus$omega;
annotation <- data.frame(
sample_id = paste0("X", c(1:NROW(omega))),
tissue_label = factor(tissue_labels[c(cerebellum_labels, hemisphere_labels)],
                      levels=c("Brain - Cerebellum", "Brain - Cerebellar Hemisphere"))
)

rownames(omega) <- annotation$sample_id;
CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))

```

## Support Vector Machine (SVM) application 

We filter out the first $50$ samples from the Blood samples and the first $50$ samples from the Pancreas tissues and then perform classification of the other samples based on these two sets of samples. 

### On counts

```{r echo=TRUE, eval=TRUE}

gene_names <- read.table("../data/GTEX_V6/gene_names_GTEX_V6.txt");
gene_names <- substring(gene_names[,1],1,15);

training_data <- t(cbind(matdata[,cerebellum_labels[1:25]],matdata[,hemisphere_labels[1:25]]))

test_data <- t(cbind(matdata[,cerebellum_labels[-(1:25)]],matdata[,hemisphere_labels[-(1:25)]]))

tissue_type <- c(rep("Cerebellum",25), rep("Cerebellar Hemisphere",25));

training.data.frame <- cbind.data.frame(training_data, tissue_type);
colnames(training.data.frame) <- c(gene_names, "tissue_type");

model <- svm(tissue_type ~ ., data = training.data.frame)

test_class_svm <- predict(model, test_data)

tab_class_svm <- table(test_class_svm, c(rep("Cerebellum", length(cerebellum_labels)-25), rep("Cerebellar Hemisphere", length(hemisphere_labels)-25)))

print(tab_class_svm)

misclass_svm <- (tab_class_svm[1,2]+tab_class_svm[2,1])/sum(tab_class_svm)
misclass_svm
```

### On voom transformed counts

```{r echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
voom_matdata <- data.frame(fread("../data/GTEX_V6/voom_cis_gene_expression.txt"))
voom_matdata <- voom_matdata[,-1]

training_data <- t(cbind(voom_matdata[,cerebellum_labels[1:25]],voom_matdata[,hemisphere_labels[1:25]]))

test_data <- t(cbind(voom_matdata[,cerebellum_labels[-(1:25)]],voom_matdata[,hemisphere_labels[-(1:25)]]))

tissue_type <- c(rep("Cerebellum",25), rep("Cerebellar Hemisphere",25));

training.data.frame <- cbind.data.frame(training_data, tissue_type);
colnames(training.data.frame) <- c(gene_names, "tissue_type");

model <- svm(tissue_type ~ ., data = training.data.frame)

test_class_svm <- predict(model, test_data)

tab_class_svm <- table(test_class_svm, c(rep("Cerebellum", length(cerebellum_labels)-25), rep("Cerebellar Hemisphere", length(hemisphere_labels)-25)))

print(tab_class_svm)

misclass_svm <- (tab_class_svm[1,2]+tab_class_svm[2,1])/sum(tab_class_svm)
misclass_svm
```

## classtpx (theta.fix with shrink using voom/lmFit) application

```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,25), rep(2,25));
known_samples <- c(1:25, length(cerebellum_labels) + (1:25));

Topic_clus <- class_topics(
    pooled_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    shrink=TRUE,
    shrink.method=1,
    tol=0.001,
    ord=FALSE)

save(Topic_clus, file="../rdas/cerebellum_hemisphere_theta_fix_classtpx_voom.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/cerebellum_hemisphere_theta_fix_classtpx_voom.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Cerebellum", "Cerebellar Hemisphere")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(rep("Cerebellum", length(cerebellum_labels)), 
                              rep("Cerebellar Hemisphere", length(hemisphere_labels)))[c(26:length(cerebellum_labels), (length(cerebellum_labels)+26):dim(omega)[1])],
                            test_class_classtpx[c(26:length(cerebellum_labels), (length(cerebellum_labels)+26):dim(omega)[1])]);

print(tab_class_classtpx)

misclass_classtpx <- (tab_class_classtpx[1,2]+tab_class_classtpx[2,1])/sum(tab_class_classtpx)
misclass_classtpx
```

## classtpx (theta.fix with shrink on counts) application

```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,25), rep(2,25));
known_samples <- c(1:25, length(cerebellum_labels) + (1:25));

Topic_clus <- classtpx::class_topics(
    pooled_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    shrink=TRUE,
    shrink.method = 2,
    tol=0.01,
    ord=FALSE)

save(Topic_clus, file="../rdas/cerebellum_hemisphere_theta_fix_classtpx.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/cerebellum_hemisphere_theta_fix_classtpx.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Cerebellum", "Cerebellar Hemisphere")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(rep("Cerebellum", length(cerebellum_labels)), 
                              rep("Cerebellar Hemisphere", length(hemisphere_labels)))[c(26:length(cerebellum_labels), (length(cerebellum_labels)+26):dim(omega)[1])],
                            test_class_classtpx[c(26:length(cerebellum_labels), (length(cerebellum_labels)+26):dim(omega)[1])]);

print(tab_class_classtpx)

misclass_classtpx <- (tab_class_classtpx[1,2]+tab_class_classtpx[2,1])/sum(tab_class_classtpx)
misclass_classtpx
```

## classtpx (theta.fix and no shrink) application

```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,25), rep(2,25));
known_samples <- c(1:25, length(cerebellum_labels) + (1:25));

Topic_clus <- classtpx::class_topics(
    pooled_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    shrink=FALSE,
    tol=0.01,
    ord=FALSE)

save(Topic_clus, file="../rdas/cerebellum_hemisphere_theta_fix_classtpx_no_shrink.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/cerebellum_hemisphere_theta_fix_classtpx_no_shrink.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Cerebellum", "Cerebellar Hemisphere")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(rep("Cerebellum", length(cerebellum_labels)), 
                              rep("Cerebellar Hemisphere", length(hemisphere_labels)))[c(26:length(cerebellum_labels), (length(cerebellum_labels)+26):dim(omega)[1])],
                            test_class_classtpx[c(26:length(cerebellum_labels), (length(cerebellum_labels)+26):dim(omega)[1])]);

print(tab_class_classtpx)

misclass_classtpx <- (tab_class_classtpx[1,2]+tab_class_classtpx[2,1])/sum(tab_class_classtpx)
misclass_classtpx
```

## classtpx (omega.fix) application

```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,25), rep(2,25));
known_samples <- c(1:25, length(cerebellum_labels) + (1:25));

Topic_clus <- classtpx::class_topics(
    pooled_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="omega.fix",
    shrink=FALSE,
    tol=0.01,
    ord=FALSE)

save(Topic_clus, file="../rdas/cerebellum_hemisphere_omega_fix_classtpx.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/cerebellum_hemisphere_omega_fix_classtpx.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Cerebellum", "Cerebellar Hemisphere")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(rep("Cerebellum", length(cerebellum_labels)), 
                              rep("Cerebellar Hemisphere", length(hemisphere_labels)))[c(26:length(cerebellum_labels), (length(cerebellum_labels)+26):dim(omega)[1])],
                            test_class_classtpx[c(26:length(cerebellum_labels), (length(cerebellum_labels)+26):dim(omega)[1])]);

print(tab_class_classtpx)

misclass_classtpx <- (tab_class_classtpx[1,2]+tab_class_classtpx[2,1])/sum(tab_class_classtpx)
misclass_classtpx
```

