---
title: 'Parametric vs Nonparametric cellcycleR: Study2'
author: "Kushal K Dey"
date: "January 25, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Overview

This script is one in a series of case studies to compare between the parametric and nonparameteric cellcycleR methods. For nonparameteric case, we use the wavelet smoother as it is likely more efficient compared to B-splines and LOESS smoothers. 

## Simulation Example

We present a non-sinusoidal gene patterns scenario now. 

```{r echo=TRUE, eval=TRUE}
library(cellcycleR)
library(wavethresh)
library(binhf)

G <- 100;
num_cells <- 256;
phi_genes <- seq(1, pi, length.out=G);
sigma_genes <- rchisq(G, 0.01);
cell_times_sim <- sort(sample(seq(0,2*pi, 2*pi/(num_cells-1)), num_cells, replace=FALSE));
cycle_data <- matrix(0,num_cells,G)
base <- c(1:(num_cells/4), (num_cells/4):1, rep(0,num_cells/4), rep(0,num_cells/4));
for(g in 1:G){
cycle_data[,g] <- rnorm(num_cells,shift(base,2*g, dir="right"),sigma_genes[g]);
}

plot(cycle_data[,1], type="l")
plot(cycle_data[,60], type="l")
plot(cycle_data[,30], type="l")


celltime_levels <- 256;

sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
cell_times_reorder <- cell_times_sim[sample_reorder];
cycle_data_reorder <- cycle_data[sample_reorder,];

```

## Sinusoidal cellcycleR

```{r echo=TRUE, eval=TRUE, cache=TRUE}
system.time(out_sinusoidal <- sin_cell_ordering_class(cycle_data_reorder, celltime_levels = 256, num_iter=500))
```

```{r echo=TRUE, eval=TRUE}
library(plotrix)
library(RColorBrewer)
radial.plot(lengths=1:length(out_sinusoidal$cell_times),radial.pos=out_sinusoidal$cell_times[order(cell_times_reorder)],
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(out_sinusoidal$cell_times)), lwd=2)

radial.plot(lengths=1:length(cell_times_reorder),radial.pos=sort(cell_times_reorder),
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(cell_times_reorder)), lwd=2)
```

The plots of estimated gene pattern and the true gene pattern.

### First co-ordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),1], type="l")
plot(cycle_data[,1],type="l")
```


### 30th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),30], type="l")
plot(cycle_data[,30],type="l")
```

### 50th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),50], type="l")
plot(cycle_data[,50],type="l")
```


## Nonparametric cellcycleR

```{r echo=TRUE, eval=TRUE, cache=TRUE}
system.time(out_np <- np_cell_ordering_class(cycle_data_reorder, celltime_levels = 256, num_iter=500))
```

```{r echo=TRUE, eval=TRUE}
library(plotrix)
library(RColorBrewer)
radial.plot(lengths=1:length(out_np$cell_times),radial.pos=out_np$cell_times[order(cell_times_reorder)],
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(out_np$cell_times)), lwd=2)

radial.plot(lengths=1:length(cell_times_reorder),radial.pos=sort(cell_times_reorder),
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(cell_times_reorder)), lwd=2)
```

The plots of estimated gene pattern and the true gene pattern.

### First co-ordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),1], type="l")
plot(cycle_data[,1],type="l")
```


### 30th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),30], type="l")
plot(cycle_data[,30],type="l")
```

### 50th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),50], type="l")
plot(cycle_data[,50],type="l")
```
