---
title: 'classtpx: Simulation design 1'
author: "Kushal K Dey"
date: "February 8, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Classtpx overview

`classtpx` is a R package meant for semi-supervised topic modeling (see [site](https://github.com/kkdey/classtpx)). It is a modification of the `maptpx` package due to Matt Taddy, which performs unsupervised topic models. As a first illustration of the application of the `classtpx` package, we present a simulation example and apply the algorithm on the simulated data. We also compare the results from the `classtpx` application with the `maptpx` method. 


## Simulation Design

Assume K = 2 hidden subpopulations in the data, and G = 600 genes. We set

$\theta_{1g} = c(\mathbb{1}_200, \mathbb{0}_200, seq(0.6, 0.4, lenght.out = 200)$

and $\sum_{g=1}^{G} \theta_{kg} = 1$.

```{r echo=TRUE, eval=TRUE}
n.out <- 200
omega_sim <- rbind( cbind( rep(1, n.out), rep(0, n.out)), 
                    cbind( rep(0, n.out), rep(1, n.out)),
                    cbind( seq(0.6, 0.4, length.out = n.out), 
                           1- seq(0.6, 0.4,length.out=n.out)) )
dim(omega_sim)
```

```{r}
K <- dim(omega_sim)[2]
barplot(t(omega_sim), 
        col = 2:(K+1),
        axisnames = F, space = 0, 
        border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim = c(0,1), cex.axis=1.5,cex.main=1.4)
```


We generate the cluster or topic distributions. These are also known as membership weights of individual cells. Here we simulate n = 100 cells, and membership weight
matrix $q_{n.}$ is

$\begin{pmatrix} q_{11} & q_{21} & \dots & q_{n1} \\
  q_{12} & q_{22} & \dots & q_{n2} 
  \end{pmatrix}$


```{r echo=TRUE, eval=TRUE}
freq <- rbind( c(0.1, 0.2, rep(0.70/98, 98)),
               c(rep(0.70/98,98), 0.1,0.2) )
str(freq)
freq[ ,1:5]
```


Now we generate the counts table given the topic proportions matrix and the topic distributions.

For every gene, we assume that $C_{n.}$ the vector of read counts coming from genes 1 to G follows a multinomial distribution where $\left( p_{n,1}, \dots, p_{n,G} \right)$ the probabilities of a read coming from genes 1 to $g$ sum up to 1. That is, 

$C_{n.} \sim Multinomial \left( C_{n+}, p_{n.} \right)$ 

and 

$p_{n.}= q^{\prime}_{n.} \begin{pmatrix} \theta_{1.} \\ \theta{2.} \end{pmatrix}$


```{r echo=TRUE, eval=TRUE}
counts <- t( do.call(cbind,
                     lapply(1:dim(omega_sim)[1], 
                            function(x) 
                                rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)
```


## classtpx Application

We assume that we have FACS sorted the cells into two known cell types. Cells numberd $1$ to $100$ are assigend to type 1, and cells $200$ to $300$ are assigned to type 2.


```{r echo=TRUE, eval=TRUE}
library(classtpx)
known_indices <- c(1:100, 201:300)
omega_known <- omega_sim[known_indices,];
Topic_clus <- classtpx::class_topics(counts, K, known_indices = known_indices, omega_known = omega_known, tol=0.0001);
```

### Structure plot 

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
docweights <- Topic_clus$omega;
library(permute);
library(BioPhysConnectoR)
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
    temp=docweights[,perm_set[p,]];
    diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

## maptpx Application

```{r echo=TRUE, eval=TRUE}
library(maptpx)
Topic_clus <- maptpx::topics(counts, K, tol=0.001);
```

### Structure plot 

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
docweights <- Topic_clus$omega;
library(permute);
library(BioPhysConnectoR)
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
    temp=docweights[,perm_set[p,]];
    diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```
