---
title: 'classtpx: Simulation Design 2'
author: "Kushal K Dey"
date: "February 8, 2016"
output: 
    html_document:
        css: floating-toc.css
        toc: true
---

## Classtpx overview

`classtpx` is a R package meant for semi-supervised topic modeling (see [site](https://github.com/kkdey/classtpx)). It is a modification of the `maptpx` package due to Matt Taddy, which performs unsupervised topic models. As a first illustration of the application of the `classtpx` package, we present a simulation example and apply the algorithm on the simulated data. We also compare the results from the `classtpx` application with the `maptpx` method. 

## Simulation Design

```{r echo=TRUE, eval=TRUE}
n.out <- 600
omega_sim <-  cbind(seq(0.6,0.4,length.out=n.out), 
                    1- seq(0.6,0.4,length.out=n.out) )
colSums(omega_sim)

K <- dim(omega_sim)[2]
barplot(t(omega_sim),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

We generate the cluster or topic distributions.

```{r echo=TRUE, eval=TRUE}
freq <- rbind( c(0.1, 0.2, rep(0.70/98,98)),
               c(rep(0.70/98,98), 0.1, 0.2))
```

Now we generate the counts table given the topic proportions matrix and the topic distributions.

```{r echo=TRUE, eval=TRUE}
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq))));
```

## classtpx Application

We assume that we have FACS sorted the cells $1$ to $100$ and from $501$ to $600$.

```{r echo=TRUE, eval=TRUE}
library(classtpx)
known_samples <- c(1:100, 501:600);
class_labs <- c(rep(1,100), rep(2,100));
```

### omega.fix method

```{r echo=TRUE, eval=TRUE}
Topic_clus <- classtpx::class_topics(
    counts, 
    K, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="omega.fix",
    tol=0.001)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
# These are estimated gene relative expression
docweights <- Topic_clus$omega

library(permute)
library(BioPhysConnectoR)

perm_set <- rbind(1:K,allPerms(1:K))
diff <- array(0,dim(perm_set)[1]) 
for (p in 1:dim(perm_set)[1])
{
    temp <- docweights[, perm_set[p,]]
    diff[p] <- fnorm(temp,omega_sim)
}

p_star <- which(diff==min(diff))
docweights <- docweights[,perm_set[p_star,]]

barplot(t(docweights),
        col = 2:(K+1),
        axisnames = F, space = 0, border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim=c(0,1), cex.axis=1.5, cex.main=1.4)
```

### theta.fix method

```{r echo=TRUE, eval=TRUE}
Topic_clus <- classtpx::class_topics(
    counts, 
    K, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    tol=0.01)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
# These are estimated gene relative expression
docweights <- Topic_clus$omega

library(permute)
library(BioPhysConnectoR)

perm_set <- rbind(1:K,allPerms(1:K))
diff <- array(0,dim(perm_set)[1]) 
for (p in 1:dim(perm_set)[1])
{
    temp <- docweights[, perm_set[p,]]
    diff[p] <- fnorm(temp,omega_sim)
}

p_star <- which(diff==min(diff))
docweights <- docweights[,perm_set[p_star,]]

barplot(t(docweights),
        col = 2:(K+1),
        axisnames = F, space = 0, border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim=c(0,1), cex.axis=1.5, cex.main=1.4)
```

### no.fix  method

```{r echo=TRUE, eval=TRUE}
Topic_clus <- classtpx::class_topics(
    counts, 
    K, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="no.fix",
    tol=0.01)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
docweights <- Topic_clus$omega;
library(permute);
library(BioPhysConnectoR)
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
    temp=docweights[,perm_set[p,]];
    diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

## maptpx application

```{r echo=TRUE, eval=TRUE}
Topic_clus <- maptpx::topics(
    counts, 
    K,
    tol=0.01)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
docweights <- Topic_clus$omega;
library(permute);
library(BioPhysConnectoR)
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
    temp=docweights[,perm_set[p,]];
    diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```
