---
title: 'classtpx: Simulation design 1'
author: "Kushal K Dey"
date: "February 8, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Classtpx overview

`classtpx` is a R package meant for semi-supervised topic modeling (see [site](https://github.com/kkdey/classtpx)). It is a modification of the `maptpx` package due to Matt Taddy, which performs unsupervised topic models. As a first illustration of the application of the `classtpx` package, we present a simulation example and apply the algorithm on the simulated data. We also compare the results from the `classtpx` application with the `maptpx` method. 

```{r}
library(classtpx)
```


## Simulation Design

Assume K = 2 hidden subpopulations in the data, and n = 600 cells. We set a 600 by 2 membership weight matrix $\mathbf{q} with $\sum_{K=2} q_{n.} = 1$, where $q_{n.}$ denotes the vector of membership weight (grade?) of cell $n$. 


```{r echo=TRUE, eval=TRUE}
n.out <- 200
omega_sim <- rbind( cbind( rep(1, n.out), rep(0, n.out)), 
                    cbind( rep(0, n.out), rep(1, n.out)),
                    cbind( seq(0.6, 0.4, length.out = n.out), 
                           1- seq(0.6, 0.4,length.out=n.out)) )
dim(omega_sim)
```


```{r}
K <- dim(omega_sim)[2]
barplot(t(omega_sim), 
        col = 2:(K+1),
        axisnames = F, space = 0, 
        border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim = c(0,1), cex.axis=1.5,cex.main=1.4)
```


We generate the cluster or topic distributions. These are also known as relative gene expression weights of individual cells. Here we simulate $G = 100$ genes, and membership weight

$\begin{pmatrix} q_{,11} & q_{1,2} & \dots & q_{1,G} \\
  q_{2,1} & q_{2,2} & \dots & q_{2,G} 
  \end{pmatrix}$


```{r echo=TRUE, eval=TRUE}
freq <- rbind( c(0.1, 0.2, rep(0.70/98, 98)),
               c(rep(0.70/98,98), 0.1, 0.2) )
str(freq)
```


Now we generate the counts table given the topic proportions matrix and the topic distributions.

For every cell $n$, we assume that $C_{n.}$ the vector of read counts coming from genes 1 to G follows a multinomial distribution where $\left( p_{n,1}, \dots, p_{n,G} \right)$ the probabilities of a read coming from genes 1 to $G$ sum up to 1. That is, 

$C_{n.} \sim Multinomial \left( C_{n+}, p_{n.} \right)$ 

and 

$p_{n.}= \sum_{k=1}^{K} q_{nk} \theta_{kg}$.


```{r echo=TRUE, eval=TRUE}
counts <- t( do.call(cbind,
                     lapply(1:dim(omega_sim)[1], 
                            function(x) 
                                rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)
```


## classtpx Application

We assume that we have FACS sorted the cells into two known cell types. Cells numberd $1$ to $100$ are assigend to type 1, and cells $200$ to $300$ are assigned to type 2.

```{r echo=TRUE, eval=TRUE}
library(classtpx)
known_samples <- c(1:100, 201:300)
class_labs <- c(rep(1,100), rep(2,100))
```


### omega.fix method

```{r echo=TRUE, eval=TRUE}
Topic_clus <- classtpx::class_topics(
    counts, 
    K, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="omega.fix",
    tol=0.001)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
# These are estimated gene relative expression
docweights <- Topic_clus$omega

library(permute)
library(BioPhysConnectoR)

# Decide on the correspondance of the simulated
# and of the estimated relative expression matrix
# 
# We consider all possible cluster ordering in the 
# estimated relative expression matrix and select
# the relative expression matrix with the nearest
# distance to the simulate relative experssion matrix
# (forbieus norm)

perm_set <- rbind(1:K,allPerms(1:K))
diff <- array(0,dim(perm_set)[1]) 
for (p in 1:dim(perm_set)[1])
{
    temp <- docweights[, perm_set[p,]]
    diff[p] <- fnorm(temp,omega_sim)
}

p_star <- which(diff==min(diff))
docweights <- docweights[,perm_set[p_star,]]

barplot(t(docweights),
        col = 2:(K+1),
        axisnames = F, space = 0, border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim=c(0,1), cex.axis=1.5, cex.main=1.4)
```

### theta.fix method

```{r echo=TRUE, eval=TRUE}
Topic_clus <- classtpx::class_topics(
    counts, 
    K, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    tol=0.01)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
# These are estimated gene relative expression
docweights <- Topic_clus$omega

library(permute)
library(BioPhysConnectoR)

# Decide on the correspondance of the simulated
# and of the estimated relative expression matrix
# 
# We consider all possible cluster ordering in the 
# estimated relative expression matrix and select
# the relative expression matrix with the nearest
# distance to the simulate relative experssion matrix
# (forbieus norm)

perm_set <- rbind(1:K,allPerms(1:K))
diff <- array(0,dim(perm_set)[1]) 
for (p in 1:dim(perm_set)[1])
{
    temp <- docweights[, perm_set[p,]]
    diff[p] <- fnorm(temp,omega_sim)
}

p_star <- which(diff==min(diff))
docweights <- docweights[,perm_set[p_star,]]

barplot(t(docweights),
        col = 2:(K+1),
        axisnames = F, space = 0, border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim=c(0,1), cex.axis=1.5, cex.main=1.4)
```

### no.fix  method

```{r echo=TRUE, eval=TRUE}
Topic_clus <- classtpx::class_topics(
    counts, 
    K, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="no.fix",
    tol=0.01)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
docweights <- Topic_clus$omega;
library(permute);
library(BioPhysConnectoR)
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
    temp=docweights[,perm_set[p,]];
    diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

## maptpx application

```{r echo=TRUE, eval=TRUE}
Topic_clus <- maptpx::topics(
    counts, 
    K,
    tol=0.01)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
docweights <- Topic_clus$omega;
library(permute);
library(BioPhysConnectoR)
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
    temp=docweights[,perm_set[p,]];
    diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```
