---
title: 'Estimate gene correlations'
author: "Joyce Hsiao"
date: "January 26, 2016"
output: 
  html_document:
    toc: true
---

## Background and objective

We assess the possibility of correlated gene expression pattern using iPSC data.


## Load packages

Load packages.

```{r echo=FALSE, eval=TRUE}
library(cellcycleR)
library(wavethresh)
```


## iPSC

Load data. Use the filtered data before any transformation.

```{r}
```

Randomly select some genes and plot out the simulated expression pattern.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data[ ,which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Simulated pattern", outer = TRUE, line = -1)
```

Shuffle labels of cell-phase (pi) between cells.

```{r echo=TRUE, eval=TRUE}
sample_reorder <- sample(1:num_cells, num_cells, replace=FALSE)
cell_times_reorder <- cell_times_sim[sample_reorder]
cycle_data_reorder <- cycle_data[sample_reorder,]

par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ ,which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
```


### Sinusoidal cellcycleR 

```{r echo=TRUE, eval=TRUE}
system.time(out <- sin_cell_ordering_class(cycle_data_reorder, 
                                           celltime_levels = 100, num_iter = 100) )
```

Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(out$cell_times), which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Sinusoidal model", outer = TRUE, line = -1)
```


### Smoothing splines fitting

```{r echo=TRUE, eval=TRUE}
system.time(
  splines_fit <- np_cell_ordering_class(cycle_data_reorder, 
                                        celltime_levels = 100, 
                                        num_iter = 100, 
                                        method = "B-spline")
  )

str(splines_fit)
```

Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(splines_fit$cell_times), which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Smoothing splines", outer = TRUE, line = -1)
```


## Case 2: Non-sinusoidal model, correlated expression

Set parameters.

```{r}
G <- 100
num_cells <- 256
phi_genes <- seq(1, pi, length.out=G)
cell_times_sim <- seq(0,2*pi, length.out=num_cells)
correlation <- .3
Sigma <- diag(G)
Sigma[upper.tri(Sigma)] <- correlation^2
Sigma[lower.tri(Sigma)] <- correlation^2

cycle_data <- matrix(0,num_cells,G)
for(s in 1:num_cells){
  	 cycle_data[s,] <- 10*cell_times_sim[s] + 30*sin(2*cell_times_sim[s] + phi_genes) + 
  	               mvrnorm(n = 1, mu = rep(0, G), Sigma = Sigma)
}
```

Shuffle labels of cell-phase (pi) between cells.

```{r echo=TRUE, eval=TRUE}
sample_reorder <- sample(1:num_cells, num_cells, replace=FALSE)
cell_times_reorder <- cell_times_sim[sample_reorder]
cycle_data_reorder <- cycle_data[sample_reorder,]

par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ ,which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
```


### Sinusoidal cellcycleR 

```{r echo=TRUE, eval=TRUE}
system.time(
  sinusoidal_fit <- sin_cell_ordering_class(cycle_data_reorder, 
                                            celltime_levels = 100, num_iter = 100) 
  )
```


Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(sinusoidal_fit$cell_times), which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Sinusoidal model", outer = TRUE, line = -1)
```


### Smoothing splines fitting

```{r echo=TRUE, eval=TRUE}
system.time(
  splines_fit <- np_cell_ordering_class(cycle_data_reorder, 
                                        celltime_levels = 100, 
                                        num_iter = 100, 
                                        method = "B-spline")
  )
```

Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(splines_fit$cell_times), which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Smoothing splines", outer = TRUE, line = -1)
```





