---
title: "LOESS in time ordering - validation check"
author: "Kushal K Dey"
date: "January 16, 2016"
output: html_document
---

## Background

The goal of this study is to apply LOESS/LOWESS smoothing (kernel smoothing) on cell-cycle or time series data. The main reasons for this validation check is to figure out how good the model fits the true ordered data compared to randomly ordered data. Also, time complexity is a big issue. The other candidate models to using LOESS would be using a spline or  wavelets. 

## Simulation from sinusoidal model

```{r, echo=TRUE, eval=TRUE}
library(cellcycleR)
library(wavethresh)
```


```{r echo=TRUE, eval=TRUE}
G <- 100;
num_cells <- 256;
amp_genes1 <- rep(10, G);
phi_genes1 <- rep(c(2,5), each=G/2);
sigma_genes1 <- rchisq(G, 4);
cell_times_sim <- sort(sample(seq(0,2*pi, 2*pi/(num_cells-1)), num_cells, replace=FALSE));
cycle_data <- t(sim_sinusoidal_cycle(G, amp_genes1, phi_genes1, sigma_genes1, cell_times_sim));

```

The sinusoidal expression pattern for $1$st gene 

```{r echo=TRUE, eval=TRUE}
plot(cycle_data[1,])
```

## Permutation of sample labels

Now we permute the labels of the times.

```{r echo=TRUE, eval=TRUE}
sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
cycle_data_reorder <- cycle_data[,sample_reorder];
```

We plot the expression pattern for $1$st gene.

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[1,])
```

## Applying LOESS smoothing 

We apply a LOESS smoothing to the data. For this, we use the *scatter.smooth()* function in \mathcal{R}. 


```{r echo=TRUE, eval=TRUE}
mu.est.true <- loess(cycle_data[1,] ~ cell_times_sim)
plot(mu.est.true$fitted, type = "l")

mu.est.reorder <- loess(cycle_data_reorder[1,] ~ cell_times_sim)
plot(mu.est.reorder$fitted, type = "l")
```

We then find the deviation of true observations against the fitted mean value from the above method.

```{r echo=TRUE, eval=TRUE}
disp_reorder <- sum((cycle_data_reorder[1,] - mu.est.reorder$y)^2)
disp_true <- sum((cycle_data[1,] - mu.est.true$y)^2)
```

Check that the dispersion of actual observations from the fitted mean curve is smaller for the true order compared to the re-ordered time series data.

## Validation of procedure

But this was for one gene only. We can find the average dispersion across all the genes.

```{r echo=TRUE, eval=TRUE}
disp.true <- array(0,G);
system.time(smash_disp_true <- mean(sapply(1:G, function(g)
                               {
                                  mu.est.true <- loess(cycle_data[g,] ~ cell_times_sim)
                                  out <- sum((cycle_data[g,] - mu.est.true$fitted)^2)
                                  return(out)
                               })))
```

Now we reorder the data and apply the same method $30$ times and plot the density of the dispersion values and check where this density function lies against the true value (corresponding to the true order) we obtained above.

```{r echo=TRUE, eval=TRUE}
library(parallel)
system.time(outdisp_set <- unlist(mclapply(1:20, function(l)
               {
                  sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
                  cycle_data_reorder <- cycle_data[,sample_reorder];
                  disp <- mean(sapply(1:G, function(g)
                               {
                                  mu.est.reord <- loess(cycle_data_reorder[g,] ~ cell_times_sim)
                                  out <- sum((cycle_data_reorder[g,] - mu.est.reord$fitted)^2)
                                  return(out)
                               }))
                  return(disp)
                }, mc.cores=detectCores())))

```

The dispersion set of values for re-ordered data (not true order) after repeating the re-ordering $20$ times 

```{r echo=TRUE, eval=TRUE}
print(outdisp_set)
```

The true value of dispersion corresponding to true ordering 

```{r echo=TRUE, eval=TRUE}
print(smash_disp_true)
```

