---
title: "low counts genes??…iPSC clustering"
author: "Kushal K Dey"
date: "July 21, 2015"
output: html_document
---


In this script, we try to figure out if some of the genes with very low counts form a cluster and would be sensible to remove these genes. The aim of this script is to build a density plot of based on the counts data available for each gene. 

We start with the dataset of single cells and genes table that was obtained after various QC checks and filtering of both cells and genes.

```{r packages, echo=FALSE, eval=TRUE}

suppressMessages(suppressWarnings(library(data.table)))
suppressMessages(suppressWarnings(library(maptpx)))
suppressMessages(suppressWarnings(library(gplots)))
suppressMessages(suppressWarnings(library(philentropy)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(edgeR)))


#setwd('/Users/kushal/Documents/Matthew Stephens Project/counts_clustering/Single_Cell_seq_Blischak/src/');
reads <- data.frame(fread('../data/reads.txt'),row.names=1);
molecules <-  data.frame(fread('../data/molecules.txt'),row.names=1);


quality_single_cells <- scan("../data/quality-single-cells.txt",
                             what = "character")

anno <- data.frame(fread('../data/annotation.txt'));

molecules <- molecules[, grepl("bulk", colnames(molecules)) |
                         colnames(molecules) %in% quality_single_cells]
anno <- anno[anno$well == "bulk" | anno$sample_id %in% quality_single_cells, ]
stopifnot(ncol(molecules) == nrow(anno),
          colnames(molecules) == anno$sample_id)

reads <- reads[, grepl("bulk", colnames(reads)) |
                         colnames(reads) %in% quality_single_cells]
stopifnot(ncol(reads) == nrow(anno),
          colnames(reads) == anno$sample_id)


expressed <- rowSums(molecules[, anno$well == "bulk"]) > 0 &
             rowSums(molecules[, anno$well != "bulk"]) > 0
molecules <- molecules[expressed, ]
#dim(molecules)

expressed <- rowSums(reads[, anno$well == "bulk"]) > 0 &
             rowSums(reads[, anno$well != "bulk"]) > 0
reads <- reads[expressed, ]
#dim(reads)

molecules_single <- molecules %>% select(-contains("bulk"))
reads_single <- reads %>% select(-contains("bulk"))

## Remove genes with max molecule numer larger than 1024

molecules_single <- molecules_single[apply(molecules_single,1,max) < 1024,];


```


We mainly focus on the molecules data because they correspond to UMI tagging. 

```{r echo=TRUE, eval=TRUE}

gene_counts_molecules <- rowSums(molecules_single);
plot(density(log(gene_counts_molecules+1)),xlab="log total reads per gene",ylab="density",
     main="Density plot of log total reads per gene (molecules)",col="red");

hist(log(gene_counts_molecules+1),xlab="log total reads per gene",ylab="frequency",
     main="Histogram of log total reads per gene (molecules)",col="gray")

gene_counts_reads <- rowSums(reads_single);
plot(density(log(gene_counts_reads+1)),xlab="log total reads per gene",ylab="density",
     main="Density plot of log total reads per gene (reads)",col="red");

hist(log(gene_counts_reads+1),xlab="log total reads per gene",ylab="frequency",
     main="Histogram of log total reads per gene (reads)",col="gray")


```

We now examine the sparsity in the dataset

```{r echo=TRUE, eval=TRUE}

gene_counts_zeros_molecules <- apply(molecules_single,1, function(x) length(which(x==0))/length(x));

plot(density(gene_counts_zeros_molecules),xlab="prop. zero reads per gene",ylab="density",
     main="Density plot of prop. of zero reads per gene (molecules)",col="red");

hist(gene_counts_zeros_molecules,xlab="prop. zero reads per gene",ylab="frequency",
     main="Histogram of prop. of zero reads per gene (molecules)",col="gray")

gene_counts_zeros_reads <- apply(reads_single,1, function(x) length(which(x==0))/length(x));


plot(density(gene_counts_zeros_reads),xlab="prop. zero reads per gene",ylab="density",
     main="Density plot of prop of zero reads per gene (reads)",col="red");

hist(gene_counts_zeros_reads,xlab="prop. zero reads per gene",ylab="frequency",
     main="Histogram of prop of  zero reads per gene (reads)",col="gray")


```
