---
title: 'Nonparametric cellcycleR: is it effective?'
author: "Kushal K Dey"
date: "January 18, 2016"
output: 
  html_document:
    toc: true
---

## Background

The idea of using nonparametric cellcycleR is to help recover the correct time order when the gene expression patterns follow some smooth patterns that cannot be modeled as sinusoidal function (so that the parameteric version of cellcycleR fails). In this script, we will consider a toy example on which we apply parametric cellcycleR and then test smoothness fit of nonparameteric cellcycleR on the data. 

## Example simulation

```{r echo=TRUE, eval=TRUE}
library(cellcycleR)
library(wavethresh)

G <- 100;
num_cells <- 256;
phi_genes <- seq(1, pi, length.out=G);
sigma_genes <- rchisq(G, 1);
cell_times_sim <- sort(sample(seq(0,2*pi, 2*pi/(num_cells-1)), num_cells, replace=FALSE));
cycle_data <- matrix(0,num_cells,G)
for(s in 1:num_cells){
cycle_data[s,] <- 10*cell_times_sim[s] + 30*sin(2*cell_times_sim[s] + phi_genes) + rnorm(G,0,sigma_genes);
}
```

We can check the patterns of a couple of genes we consider.

```{r echo=TRUE, eval=TRUE}
plot(cycle_data[,1], type="l")
plot(cycle_data[,60], type="l")
plot(cycle_data[,30], type="l")
```

## A cell re-ordering of the data

```{r echo=TRUE, eval=TRUE}
sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
cell_times_reorder <- cell_times_sim[sample_reorder];
cycle_data_reorder <- cycle_data[sample_reorder,];

plot(cycle_data_reorder[,1], type="l")
plot(cycle_data_reorder[,60], type="l")
plot(cycle_data_reorder[,30], type="l")

```

## Sinusoidal cellcycleR 

```{r echo=TRUE, eval=TRUE}
system.time(out <- sin_cell_ordering_class(cycle_data_reorder, celltime_levels = 128, num_iter=300))
```

We check how the patterns look like for the genes considered above under the re-ordered cell times.

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out$cell_times),1], type="l");
plot(cycle_data_reorder[order(out$cell_times),60], type="l");
plot(cycle_data_reorder[order(out$cell_times),30], type="l");
```

## Nonparameteric smoother fit on re-ordered data from sinusoidal cellcycleR

We now fit nonparameteric model fit to these patterns we recover and compare that with nonparameteric model fit on the true data. We have chosen the Wavelet method for the comparison.

```{r echo=TRUE, eval=TRUE}
system.time(smash_disp_est <- mean(sapply(1:G, function(g)
                               {
                                  mu.est.true <- smooth.spline(sort(out$cell_times), cycle_data_reorder[order(out$cell_times),g]);
                                  ordered_vec <- as.numeric(tapply(cycle_data_reorder[order(out$cell_times),g], factor(sort(out$cell_times)), mean));
                                   disp <- sum((ordered_vec - mu.est.true$y)^2)
                                  return(disp)
                               })))

print(smash_disp_est)

```

## Nonparameteric smoother fit on true data ordering 

Now we perform the same analysis on the true data with the true order.

```{r echo=TRUE, eval=TRUE}
system.time(smash_disp_true <- mean(sapply(1:G, function(g)
                               {
                                  mu.est.true <- smooth.spline(cell_times_sim, cycle_data[,g])
                                  out <- sum((cycle_data[,g] - mu.est.true$y)^2)
                                  return(out)
                               })))

print(smash_disp_true)

```

## Discussion

Note that the average squared distance between data and the fitted curve is way smaller for the re-ordered data than the true data case. One reason could be the fact that the cell times in the true data are `r length(unique(cell_times_sim)) ` while after we put them to time classes as per our cellcycleR method, we get only `r length(unique(out$cell_times))` classes based on which the model fit is compared. There is a disparity there, which may sustain despite the fact that we are taking average of the distances.

The other problem could be the fact that the squared distance of residuals may not be a good measure. In which case, we may want to have a method that under this above set up, gives a smaller value under true order than the order fitted by the parameteric cellcycleR method. Only when that happens, nonparameteric cellcycleR would be meaningful.



