---
title: 'Nonparametric cellcycleR: Comparison of methods'
author: "Kushal K Dey"
date: "January 26, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Overview

This script compares the different methods of nonparameteric cellcycleR methods. The aim of the script is to show that the different nonparametric smoothers perform almost equally well as far as the cellcycle ordering is concerned. We consider the same example as in Case Study 4.

## Simulation Example

We present a non-sinusoidal gene patterns scenario now. 

```{r echo=TRUE, eval=TRUE, cache=TRUE}
library(cellcycleR)
library(wavethresh)
library(binhf)

G <- 100;
num_cells <- 256;
phi_genes <- seq(1, pi, length.out=G);
sigma_genes <- rchisq(G, 0.01);
cell_times_sim <- sort(sample(seq(0,2*pi, 2*pi/(num_cells-1)), num_cells, replace=FALSE));
cycle_data <- matrix(0,num_cells,G)
base <- c(1:(num_cells/4), (num_cells/4):1, 0.3*(1:(num_cells/4))^2, 0.3*((num_cells/4):1)^2);
for(g in 1:G){
cycle_data[,g] <- rnorm(num_cells,shift(base,2*g, dir="right"),sigma_genes[g]);
}

plot(cycle_data[,1], type="l")
plot(cycle_data[,60], type="l")
plot(cycle_data[,30], type="l")


celltime_levels <- 256;

sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
cell_times_reorder <- cell_times_sim[sample_reorder];
cycle_data_reorder <- cycle_data[sample_reorder,];

```


## Nonparametric cellcycleR : LOESS method


```{r echo=TRUE, eval=TRUE, cache=TRUE}
system.time(out_np <- np_cell_ordering_class(cycle_data_reorder, celltime_levels = 256, num_iter=500, method="LOESS"))
```

```{r echo=TRUE, eval=TRUE}
library(plotrix)
library(RColorBrewer)
radial.plot(lengths=1:length(out_np$cell_times),radial.pos=out_np$cell_times[order(cell_times_reorder)],
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(out_np$cell_times)), lwd=2)

radial.plot(lengths=1:length(cell_times_reorder),radial.pos=sort(cell_times_reorder),
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(cell_times_reorder)), lwd=2)
```

The plots of estimated gene pattern and the true gene pattern.

### First co-ordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),1], type="l")
plot(cycle_data[,1],type="l")
```


### 30th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),30], type="l")
plot(cycle_data[,30],type="l")
```

### 50th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),50], type="l")
plot(cycle_data[,50],type="l")
```

## Nonparametric cellcycleR : B-spline method


```{r echo=TRUE, eval=TRUE, cache=TRUE}
system.time(out_np <- np_cell_ordering_class(cycle_data_reorder, celltime_levels = 256, num_iter=500, method="B-spline"))
```

```{r echo=TRUE, eval=TRUE}
library(plotrix)
library(RColorBrewer)
radial.plot(lengths=1:length(out_np$cell_times),radial.pos=out_np$cell_times[order(cell_times_reorder)],
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(out_np$cell_times)), lwd=2)

radial.plot(lengths=1:length(cell_times_reorder),radial.pos=sort(cell_times_reorder),
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(cell_times_reorder)), lwd=2)
```

The plots of estimated gene pattern and the true gene pattern.

### First co-ordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),1], type="l")
plot(cycle_data[,1],type="l")
```


### 30th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),30], type="l")
plot(cycle_data[,30],type="l")
```

### 50th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),50], type="l")
plot(cycle_data[,50],type="l")
```

## Nonparametric cellcycleR : Wavelet method


```{r echo=TRUE, eval=TRUE, cache=TRUE}
system.time(out_np <- np_cell_ordering_class(cycle_data_reorder, celltime_levels = 256, num_iter=500, method="Wavelet"))
```

```{r echo=TRUE, eval=TRUE}
library(plotrix)
library(RColorBrewer)
radial.plot(lengths=1:length(out_np$cell_times),radial.pos=out_np$cell_times[order(cell_times_reorder)],
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(out_np$cell_times)), lwd=2)

radial.plot(lengths=1:length(cell_times_reorder),radial.pos=sort(cell_times_reorder),
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(cell_times_reorder)), lwd=2)
```

The plots of estimated gene pattern and the true gene pattern.

### First co-ordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),1], type="l")
plot(cycle_data[,1],type="l")
```


### 30th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),30], type="l")
plot(cycle_data[,30],type="l")
```

### 50th coordinate

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_np$cell_times),50], type="l")
plot(cycle_data[,50],type="l")
```

