---
title: "PCA vs SNR"
author: "Kushal K Dey"
date: "January 26, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Overview

This script is aimed at showing how the Principal Components Analysis (PCA) behaves for sinusoidal patterned data. The idea is that if the gene patterns have high signal to noise variation, then the PC1 vs PC2 plot would be close to a circle. The more we add noise and randomness to the sinusoidal patterns, the less obvious the circular pattern of the PCs would tend to be.

## Simulation Example 

We present a non-sinusoidal gene patterns scenario now. 

```{r echo=TRUE, eval=TRUE, cache=TRUE}

library(cellcycleR)
library(wavethresh)

G <- 100;
num_cells <- 256;
amp_genes <- rep(10, G);
phi_genes <- rep(c(2,4,6,8), each=G/4);
sigma_genes <- rchisq(G, 0.01);
cell_times_sim <- seq(0,2*pi, length.out=num_cells);
cycle_data <- sim_sinusoidal_cycle(G, amp_genes, phi_genes, sigma_genes, cell_times_sim);

celltime_levels <- 256;

sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
cell_times_reorder <- cell_times_sim[sample_reorder];
cycle_data_reorder <- cycle_data[sample_reorder,];

```

We perform PCA on the reordered data.

```{r echo=TRUE, eval=TRUE}
pca_cycle <- prcomp(cycle_data_reorder, center=TRUE, scale. = TRUE);
plot(pca_cycle$x[,1], pca_cycle$x[,2], pch=20, lwd=0.01)
```

Note the PCA maps the points on the circle. We now apply sinusoidal cellcycleR on the data.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
system.time(out_sinusoidal <- sin_cell_ordering_class(cycle_data_reorder, celltime_levels = 256, num_iter=100))
```

We plot the radial plots first.

```{r echo=TRUE, eval=TRUE}
library(plotrix)
library(RColorBrewer)
radial.plot(lengths=1:length(out_sinusoidal$cell_times),radial.pos=out_sinusoidal$cell_times[order(cell_times_reorder)],
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(out_sinusoidal$cell_times)), lwd=2)
```

We plot the patterns for two genes (for arbitrary indices)

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),1], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data[,1], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),60], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data[,60], type="l", xlab="cell times", ylab="gene expression")
```

Now we rank the cells based on the cell times estimated from sinusoidal cellcycleR and then plot the ranks against the ranks based on the first PC. 

```{r echo=TRUE, eval=TRUE}
plot(rank(out_sinusoidal$cell_times), rank(pca_cycle$x[,1]), xlab="sin. cellcycleR rank", ylab="pca 1 rank", pch=20, lwd=0.5, cex=1, col="red")
plot(rank(out_sinusoidal$cell_times), rank(pca_cycle$x[,2]), xlab="sin. cellcycleR rank", ylab="pca 2 rank", pch=20, lwd=0.5, cex=1, col="red")
```

## Simulation example 2

```{r echo=TRUE, eval=TRUE, cache=TRUE}

library(cellcycleR)
library(wavethresh)

G <- 100;
num_cells <- 256;
amp_genes <- rep(1, G);
phi_genes <- rep(c(2,4,6,8), each=G/4);
sigma_genes <- rchisq(G, 3);
cell_times_sim <- seq(0,2*pi, length.out=num_cells);
cycle_data <- sim_sinusoidal_cycle(G, amp_genes, phi_genes, sigma_genes, cell_times_sim);

celltime_levels <- 256;

sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
cell_times_reorder <- cell_times_sim[sample_reorder];
cycle_data_reorder <- cycle_data[sample_reorder,];

```

We perform PCA on the reordered data.

```{r echo=TRUE, eval=TRUE}
pca_cycle <- prcomp(cycle_data_reorder, center=TRUE, scale. = TRUE);
plot(pca_cycle$x[,1], pca_cycle$x[,2], pch=20, lwd=0.01)
```

We apply sinusoidal cellcycleR

```{r echo=TRUE, eval=TRUE, cache=TRUE}
system.time(out_sinusoidal <- sin_cell_ordering_class(cycle_data_reorder, celltime_levels = 256, num_iter=100))
```

We plot the radial plots first.

```{r echo=TRUE, eval=TRUE}
library(plotrix)
library(RColorBrewer)
radial.plot(lengths=1:length(out_sinusoidal$cell_times),radial.pos=out_sinusoidal$cell_times[order(cell_times_reorder)],
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(out_sinusoidal$cell_times)), lwd=2)
```

We plot the patterns for two genes (for arbitrary indices)

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),1], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data[,1], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),60], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data[,60], type="l", xlab="cell times", ylab="gene expression")

```

Now we rank the cells based on the cell times estimated from sinusoidal cellcycleR and then plot the ranks against the ranks based on the first PC. 

```{r echo=TRUE, eval=TRUE}
plot(rank(out_sinusoidal$cell_times), rank(pca_cycle$x[,1]), xlab="sin. cellcycleR rank", ylab="pca 1 rank", pch=20, lwd=0.5, cex=1, col="red")
plot(rank(out_sinusoidal$cell_times), rank(pca_cycle$x[,2]), xlab="sin. cellcycleR rank", ylab="pca 2 rank", pch=20, lwd=0.5, cex=1, col="red")
```

## Simulation example 3

```{r echo=TRUE, eval=TRUE, cache=TRUE}

library(cellcycleR)
library(wavethresh)

G <- 100;
num_cells <- 256;
amp_genes <- rep(1, G);
phi_genes <- rep(c(2,4,6,8), each=G/4);
sigma_genes <- rchisq(G, 6);
cell_times_sim <- seq(0,2*pi, length.out=num_cells);
cycle_data <- sim_sinusoidal_cycle(G, amp_genes, phi_genes, sigma_genes, cell_times_sim);

celltime_levels <- 256;

sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
cell_times_reorder <- cell_times_sim[sample_reorder];
cycle_data_reorder <- cycle_data[sample_reorder,];

```

We perform PCA on the reordered data.

```{r echo=TRUE, eval=TRUE}
pca_cycle <- prcomp(cycle_data_reorder, center=TRUE, scale. = TRUE);
plot(pca_cycle$x[,1], pca_cycle$x[,2], pch=20, lwd=0.01)
```

We apply sinusoidal cellcycleR

```{r echo=TRUE, eval=TRUE, cache=TRUE}
system.time(out_sinusoidal <- sin_cell_ordering_class(cycle_data_reorder, celltime_levels = 256, num_iter=100))
```

We plot the radial plots first.

```{r echo=TRUE, eval=TRUE}
library(plotrix)
library(RColorBrewer)
radial.plot(lengths=1:length(out_sinusoidal$cell_times),radial.pos=out_sinusoidal$cell_times[order(cell_times_reorder)],
            line.col=colorRampPalette(brewer.pal(9,"Blues"))(length(out_sinusoidal$cell_times)), lwd=2)
```

We plot the patterns for two genes (for arbitrary indices)

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),1], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data[,1], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data_reorder[order(out_sinusoidal$cell_times),60], type="l", xlab="cell times", ylab="gene expression")
plot(cycle_data[,60], type="l", xlab="cell times", ylab="gene expression")

```

Now we rank the cells based on the cell times estimated from sinusoidal cellcycleR and then plot the ranks against the ranks based on the first PC. 

```{r echo=TRUE, eval=TRUE}
plot(rank(out_sinusoidal$cell_times), rank(pca_cycle$x[,1]), xlab="sin. cellcycleR rank", ylab="pca 1 rank", pch=20, lwd=0.5, cex=1, col="red")
plot(rank(out_sinusoidal$cell_times), rank(pca_cycle$x[,2]), xlab="sin. cellcycleR rank", ylab="pca 2 rank", pch=20, lwd=0.5, cex=1, col="red")
```
