---
title: "scRNASeq data sets"
author: "Joyce Hsiao"
date: "2016-03-18"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---


```{r, include = FALSE}
library(knitr)
opts_chunk$set(eval = FALSE, echo = TRUE)
```


This document list some single-cell RNA-seq data sets downloaded from GEO. We store each of the data sets in `ExpressionSet` object and make the data available as an R package on GitHub. 



## Human

* [Tung et al., 2016](https://github.com/jhsiao999/singleCellRNASeqHumanTungiPSC): Human iPSC, counts, UMI

* [Leng et al., 2015](https://github.com/jhsiao999/singleCellRNASeqHumanLengESC): Human ESC, reads, cell-cycle sorted



## Mouse

* [Buettner et al., 2015](https://github.com/jhsiao999/singleCellRNASeqMouseBuettnerESC): mouse ESC, reads, ERCC spike-in control

* [Jaitin et at., 2014](https://github.com/jhsiao999/singleCellRNASeqMouseJaitinSpleen): mouse spleen cells, counts, UMI

* [Treutlein et al., 2014](https://github.com/jhsiao999/singleCellRNASeqMouseTreutleinLung): mouse lung cells, reads

* [Zeisel et al., 2015](https://github.com/jhsiao999/singleCellRNASeqMouseZeiselBrain): Mouse brain, counts + UMI



## Make ExpressionSet packages

See this page for how we make these packages: scRNA-expressionSet.html

See this page for data description: data-description.html

## Session information

```{r}
sessionInfo()
```

