---
title: "Store datasets in ExpressionSet objects"
author: "Joyce Hsiao"
date: "2016-03-22"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---


```{r, include = FALSE}
library(knitr)
opts_chunk$set(eval = FALSE, echo = TRUE)
```


# Background and objectives

This document contains the code for storing scRNA-seq data set in `ExpressionSet` objects. Each scRNA-seq data set is stored in its own GitHub repository and can be loaded as an R package.


# Load packages

```{r}
library(data.table)
library(Biobase)
```


# single-cell RNA-seq


## Buettner et al., 2015, mouse ESC, reads + ERCC


```{r, eval = FALSE}
# import count matrix, convert to a matrix object
G1_cells <- data.frame( 
    fread('../data/MouseBuettnerESC/G1_singlecells_counts.txt'),
    row.names = 1)
G2M_cells <- data.frame(
    fread('../data/MouseBuettnerESC/G2M_singlecells_counts.txt'),
    row.names = 1)
S_cells <- data.frame(
    fread('../data/MouseBuettnerESC/S_singlecells_counts.txt'),
    row.names = 1)
stopifnot(all.equal(rownames(G1_cells), rownames(G2M_cells)))


# make a count matrix
counts <- cbind(G1_cells[, -c(1:3)],
                G2M_cells[, -c(1:3)],
                S_cells[, -c(1:3)] )

# feature info
gene_info <- data.frame(G1_cells[ ,1:3])
row_data <- gene_info

# phenotype info
meta_list <- data.frame(labelDescription = c("cell-cycle phase"),
                    row.names = c("cell_cycle"))
col_data <- data.frame(cell_cycle = as.character(
    rep(c("G1", "G2M", "S"), each = 96)) )
rownames(col_data) <- colnames(counts)
stopifnot(all.equal(rownames(col_data), colnames(counts)))
    

# putting all together
MouseBuettnerESC <- new("ExpressionSet",
                exprs = as.matrix(counts),
                phenoData = new("AnnotatedDataFrame",
                                data = col_data,
                                varMetadata = meta_list),
                featureData = new("AnnotatedDataFrame",
                                  data = row_data),
                experimentData = new("MIAME",
                                     title = "Buttner Mouse ESC"))


save(MouseBuettnerESC,
        file = "../data/all-rds/MouseBuettnerESC.rda")
```




## Jaitin et al., 2014, mouse spleen cells, counts + UMI


```{r, eval = FALSE}
# import count matrix, convert to a matrix object
counts <- as.matrix(
    read.table("../data/MouseJaitinSpleen/GSE54006_umitab.txt",
               header = TRUE, stringsAsFactors = FALSE,
               sep = "\t") )
gene_names <- counts[ ,1]
sample_names <- colnames(counts)[-1]
counts <- counts[ ,-1]
counts <- matrix(as.numeric(counts), byrow = FALSE,
                 nrow = length(gene_names))
rownames(counts) <- gene_names

# feature information
row_data <- data.frame(gene_names = gene_names)

# import annotation matrix
col_data <- read.table("../data/MouseJaitinSpleen/GSE54006_experimental_design.txt",
               header = TRUE, 
               stringsAsFactors = FALSE,
               sep = "\t") 
colnames(col_data) <- unlist(col_data[6,])
col_data <- col_data[ - c(1:6), ]
head(col_data)


# match annotation matrix with the count matrix
temp_colnames <- do.call(c,
        lapply(sample_names, function(x) strsplit(x, "X")[[1]][[2]]) )
col_data <- col_data[which(col_data$Column_name_in_processed_data_file %in%
                         temp_colnames), ]
col_data <- data.frame(col_data)

meta_list <- data.frame(row.names = colnames(col_data))

# putting all together
MouseJaitinSpleen <- new("ExpressionSet",
                exprs = as.matrix(counts),
                phenoData = new("AnnotatedDataFrame",
                                data = col_data,
                                varMetadata = meta_list),
                featureData = new("AnnotatedDataFrame",
                                  data = row_data),
                experimentData = new("MIAME",
                                     title = "Jaitin Mouse Spleen"))

save(MouseJaitinSpleen,
        file = "../data/all-rds/MouseJaitinSpleen.rda")
```






## Tung et al., 2016, Human iPSC + UMI

Import molecule counts after QC-filtering.

```{r, eval = FALSE}
# import count matrix, convert to a matrix object
counts <- as.matrix(
    read.table("../data/HumanTungiPSC/molecules-filter.txt",
               header = TRUE, stringsAsFactors = FALSE) )

# import annotation matrix, conver to a DatFrame object
col_data <- data.frame(
    read.table("../data/HumanTungiPSC/annotation-filter.txt",
               header = TRUE, stringsAsFactors = FALSE) )
rownames(col_data) <- colnames(counts)
meta_list <- data.frame(labelDescription = c("Individual cell line",
                                             "C1 replicate",
                                             "well ID",
                                             "processing batch",
                                             "unique sample ID"),
                            row.names = colnames(col_data) )

# feature data
row_data <- data.frame(gene_name = as.character(rownames(counts)),
                       stringsAsFactors = FALSE)
rownames(row_data) <- rownames(counts)

# putting all together
HumanTungiPSC <- new("ExpressionSet",
                    exprs = as.matrix(counts),
                    phenoData = new("AnnotatedDataFrame",
                                    data = col_data,
                                    varMetadata = meta_list),
                    featureData = new("AnnotatedDataFrame",
                                      data = row_data),
                    experimentData = new("MIAME",
                                         title = "Tung Human iPSC"))


save(HumanTungiPSC,
     file = "../data/all-rda/HumanTungiPSC.rda")
```


## Leng2015, reads

Two sets of experiments were performed in Leng2015. In the first set of experiments, human embryonic stem cells (hESC) were sorted into three C1 plates corresponding to cell-cycle phase of G1, S, and G2. Fluorescent ubiquitination-based cell-cycle indicator (FUCCI) was used. The H1-FUCCI cell line provides a two-color fluorescence labeling system allowing single-cell suspensions from G1, S or G2/M cell-cycle phases to be isolated by FACS. In the second set of experiments, in order to validate their method of identifying oscillating genes, they collected hESCs with unknown cell-cycles in three C1 plates.


```{r, eval = F}
# import read count matrix
leng_data <- read.csv("../data/HumanLengESC/GSE64016_H1andFUCCI_normalized_EC.csv")
rownames(leng_data) <- leng_data[ ,1]
counts <- as.matrix(leng_data[,-1])
dim(counts)


# construct annotation matrix
col_data <- colnames(counts)
col_data <- data.frame(
    cell_state = sapply(col_data, 
                       function(xx) strsplit(xx, "_")[[1]][1]),
    replicate = sapply(col_data, 
                   function(xx) strsplit(xx, "_")[[1]][2]),
    cell_id = sapply(col_data,
                     function(xx) strsplit(xx, ".", fixed = TRUE)[[1]][2]),
    sample_id = col_data)
rownames(col_data) <- colnames(counts)

meta_list <- data.frame(labelDescription = c("cell-cycle state",
                                             "C1 replicate",
                                             "Unique cell ID for each replicate",
                                             "Unique cell ID"),
                    row.names = colnames(col_data))

# make featureData
row_data <- data.frame(gene_names = rownames(counts),
                       stringsAsFactors = FALSE)
rownames(row_data) <- rownames(counts)


# putting all together
HumanLengESC <- new("ExpressionSet",
                    exprs = as.matrix(counts),
                    phenoData = new("AnnotatedDataFrame",
                                    data = col_data,
                                    varMetadata = meta_list),
                    featureData = new("AnnotatedDataFrame",
                                      data = row_data),
                    experimentData = new("MIAME",
                                         title = "Buttner Mouse ESC"))

save(HumanLengESC,
        file = "../data/all-rda/HumanLengESC.rda")
```






## Treutlein et al., 2014, Mouse lung + reads

__GSE52583__


```{r, eval = FALSE}
###<-- import count data sts
data_list <- list.files(path = "../data/MouseTreutleinLung/",
                        pattern = "*.fpkm_tracking",
                        full.name = TRUE)

data_list_gse_label <- sapply(lapply(strsplit(data_list, split = "/"), "[[",5),
                    function(x) strsplit(x, split = "_", fixed = TRUE)[[1]][1])
length(data_list_gse_label)

library(data.table)
library(dplyr)
counts_list <- lapply(1:length(data_list), function(i) {
    df <- read.table(data_list[[i]], stringsAsFactors = FALSE)[,c(1,10)]
    df <- data.table(df, key = "V1")
    df <- summarise(group_by(df, V1 ), V10 = sum( V10 ) ) 
    colnames(df) <- c("gene_id", paste0("sam_",i))
    return(df)
})

counts <- Reduce(merge, counts_list)

counts <- data.frame(counts)
rownames(counts) <- counts[,1]
counts <- counts[,-1]
colnames(counts) <- data_list_gse_label

###<--- construct annotation matrix
library(GEOquery)
gpl13112 <- getGEO(filename = "../data/MouseTreutleinLung/GSE52583-GPL13112_series_matrix.txt")
gpl16417 <- getGEO(filename = "../data/MouseTreutleinLung/GSE52583-GPL16417_series_matrix.txt")
dim(gpl13112)
dim(gpl16417)

df_col_data <- rbind(
    pData(gpl13112)[ , which(colnames(pData(gpl13112)) %in% 
                      c("title", "geo_accession", "instrument_model"))],
    pData(gpl16417)[ , which(colnames(pData(gpl16417))  %in% 
                      c("title", "geo_accession", "instrument_model"))])
cell_type <- sapply(1:length(df_col_data$title), function(i) {
    gsub(",", replacement = "",
         strsplit(as.character(df_col_data$title[i]), 
                  split = " ", fixed = TRUE)[[1]][1]) }) 
replicate <- sapply(1:length(df_col_data$title), function(i) {
    gsub(",", replacement = "",
         strsplit(as.character(df_col_data$title[i]), 
                  split = " ", fixed = TRUE)[[1]][4]) }) 

col_data <- data.frame(
    cell_type = unlist(cell_type),
    replicate = unlist(replicate),
    gse = as.character(df_col_data$geo_accession),
    unique_id = c(1:length(cell_type)),
    sample_id = paste0(cell_type,".rep", replicate, ".", c(1:length(cell_type))),
    stringsAsFactors = FALSE)
rownames(col_data) <- col_data$sample_id

meta_list <- data.frame(labelDescription = c("cell type",
                                             "biological replicate",
                                             "GEO accession number",
                                             "Unique ID",
                                             "Unique sample ID"),
                    row.names = colnames(col_data))

# match the order of the count data and the annotation data
ii_match <- match(colnames(counts), col_data$gse)
col_data <- col_data[ii_match, ]
stopifnot(all.equal(colnames(counts), col_data$gse))

# rename columns of the count matrix to unique sample ID
colnames(counts) <- col_data$sample_id

###<--- make featureData
row_data <- data.frame(gene_names = rownames(counts),
                       stringsAsFactors = FALSE)
rownames(row_data) <- rownames(counts)


# putting all together
MouseTreutleinLung <- new("ExpressionSet",
                    exprs = as.matrix(counts),
                    phenoData = new("AnnotatedDataFrame",
                                    data = col_data,
                                    varMetadata = meta_list),
                    featureData = new("AnnotatedDataFrame",
                                      data = row_data),
                    experimentData = new("MIAME",
                                         title = "Treutlein mouse lung"))

save(MouseTreutleinLung,
        file = "../data/all-rda/MouseTreutleinLung.rda")
```






## Ziesel et al., 2015, Mouse brain + UMI

http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE60361

> We have applied a recently developed, highly accurate and sensitive single-cell RNA-seq method (STRT/C1) to perform a molecular census of two regions of the mouse cerebral cortex: the somatosensory cortex and hippocampus CA1. We isolated cells fresh from somatosensory cortex (S1) and hippocampus CA1 area of juvenile (P22 - P32) CD1 mice, 33 males and 34 females. Cells were collected without selection, except that 116 cells were obtained by FACS from 5HT3a-BACEGFP transgenic mice. A total of 76 Fluidigm C1 runs were performed, each attempting 96 cell captures and resulting in 3005 high-quality single-cell cDNAs, containing Unique Molecular Identifiers allowing counting of individual mRNA molecules, even after PCR amplification.


```{r, eval = F}
library(data.table)
library(Biobase)

# import annotation matrix, conver to a DatFrame object
df_col_data <- as.matrix(
    fread("../data/MouseZeiselBrain/expression_mRNA_17-Aug-2014.txt") )[1:10, ]

col_data <- data.frame(
    tissue = colnames(df_col_data)[-(1:2)],
    group_no = round(as.numeric(df_col_data[1, -(1:2)])),
    total_mrna_vol = as.numeric(df_col_data[2, -(1:2)]),
    well_id = round(as.numeric(df_col_data[3, -(1:2)])),
    sex = factor(as.numeric(df_col_data[4, -(1:2)]),
                 levels = c(-1, 1, 0),
                 labels = c("male", "female", "transgenic")),
    age = as.numeric(df_col_data[5, -(1:2)]),
    diameter = as.numeric(df_col_data[6, -(1:2)]),
    cell_id = df_col_data[7, -(1:2)],
    level1_class = df_col_data[8, -(1:2)],
    level2_class = df_col_data[9, -(1:2)],
    stringsAsFactors = FALSE)
rownames(col_data) <- col_data$cell_id



# import data
df <- as.matrix(
    read.table("../data/MouseZeiselBrain/GSE60361_C1-3005-Expression.txt",
    header = TRUE, quote = "", sep = "\t",
    stringsAsFactors = FALSE))

# save counts to a matrix object
counts <- matrix(as.numeric(df[ ,-1]),
                    nrow = NROW(df),
                    byrow = FALSE)
colnames(counts) <- colnames(df)[-1]
rownames(counts) <- df[,1]


# two genes are duplicates...
rownames(counts)[which(duplicated(rownames(counts)))]

# check the difference between the duplicates
table((counts[rownames(counts) == "Mar-02", ])[1,])
table((counts[rownames(counts) == "Mar-02", ])[2,])

# check the difference between the duplicates
table((counts[rownames(counts) == "Mar-01", ])[1,])
table((counts[rownames(counts) == "Mar-01", ])[2,])

# removing these two genes as I am not sure the duplicates
# are drastically different in gene abundance levels
counts <- counts[which(rownames(counts) != "Mar-01" & rownames(counts) != "Mar-02"), ]

# remote X in the beginning of column names
counts_colnames <- gsub("X", replacement = "", x = colnames(counts))
colnames(counts) <- counts_colnames

# feature data
row_data <- data.frame(gene_name = as.character(rownames(counts)),
                       stringsAsFactors = FALSE)
rownames(row_data) <- rownames(counts)

# match count matrix with annotation data.frame
match_cell_id <- match(col_data$cell_id, colnames(counts))
counts <- counts[ ,match_cell_id]

meta_list <- data.frame(
    labelDescription = c("Tissue type",
                         "Group nubmer",
                         "Total mRNA volume",
                         "Well ID",
                         "Sex",
                         "Age (in days)",
                         "Cell diameter",
                         "Cell unique ID",
                         "Level 1 class",
                         "Level 2 class"),
    row.names = colnames(col_data) )


# putting all together
MouseZeiselBrain <- new("ExpressionSet",
                        exprs = as.matrix(counts),
                        phenoData = new("AnnotatedDataFrame",
                                        data = col_data,
                                        varMetadata = meta_list),
                        featureData = new("AnnotatedDataFrame",
                                          data = row_data),
                        experimentData = new("MIAME",
                                             title = "Zeisel Mouse Brain"))


save(MouseZeiselBrain,
     file = "../data/all-rda/MouseZeiselBrain.rda")
```


## Deng et al 2014, counts

```{r, eval = F}
reads <- data.frame(data.table::fread('../data/Deng_Data/Deng_cell_data.txt'),row.names=1);
files <- list.files("../data/Deng_Data/Deng_files/");

cell_meta <- unlist(lapply(files, function(x) strsplit(x,"_")[[1]][2]));
embryo_id <- unlist(lapply(files, function(x) strsplit(strsplit(x,"_")[[1]][3],"-")[[1]][1]));
cell_meta[grep("zy",cell_meta)]="zy";
cell_meta[grep("smartseq2", files)]="8cell_nd";
cell_meta[grep("8cell_2pooled", files)]="8cell_nd";
cell_meta[grep("8cell_split", files)]="8cell_nd";
cell_meta[grep("16cell_2pooled", files)]="16cell_nd";
cell_meta[grep("16cell_split", files)]="16cell_nd";
indices_not_reqd <- which(cell_meta=="BXC"   | cell_meta=="C57twocell" | cell_meta=="fibroblast" | cell_meta =="8cell_nd" | cell_meta == "16cell_nd");
cell_meta <- cell_meta[-indices_not_reqd];
embryo_id <- embryo_id[-indices_not_reqd];
embryo_id[which(embryo_id == "expression.txt")]="."

reads <- reads[,-indices_not_reqd];
cell_meta_unique <- c("zy","early2cell","mid2cell","late2cell","4cell","8cell","16cell","earlyblast","midblast","lateblast") ;
order_of_development <- order(match(cell_meta,cell_meta_unique))
reads <- reads[,order_of_development];
cell_meta <- cell_meta[order_of_development]
embryo_id <- embryo_id[order_of_development]

row_data <- data.frame(gene_name = as.character(rownames(reads)),
                       stringsAsFactors = FALSE)
rownames(row_data) <- rownames(reads)

col_data <- data.frame(
    cell_type = cell_meta,
    embryo_id = embryo_id);

colnames(col_data) <- c("cell_type", "embryo_id");
rownames(col_data) <- colnames(reads);

meta_list <- data.frame(labelDescription = c("cell-type",
                                             "embryo-id"),
                    row.names = colnames(col_data))

Deng2014MouseESC <- new("ExpressionSet",
                exprs = as.matrix(reads),
                phenoData = new("AnnotatedDataFrame",
                                data = col_data),
                featureData = new("AnnotatedDataFrame",
                                  data = row_data),
                experimentData = new("MIAME",
                                     title = "Deng 2014 Mouse ESC"))

save(Deng2014MouseESC,
        file = "../data/all-rds/Deng2014MouseEsc.rda")
```

<<<<<<< HEAD
## Scialdone et al 2015, reads

```{r}
library(data.table)
counts_liver <- data.frame(
    fread("../data/MouseScialdoneLiver/Marioni_lab_1_Jul_2015_gene_counts_table.txt"))
dim(counts_liver)
str(counts_liver)

row_data <- data.frame(gene_name = as.character(counts_liver$V1),
                       stringsAsFactors = FALSE)
rownames(row_data) <- counts_liver$V1

col_data <- data.frame(sample_id = paste0("sampleID",c(1:96)))
rownames(col_data) <- paste0("sampleID",c(1:96))

meta_list <- data.frame(labelDescription = c("sampleID"),
                    row.names = colnames(col_data))

counts_liver <- counts_liver[,-1]
rownames(counts_liver) <- row_data$gene_name
colnames(counts_liver) <- col_data$sample_id

# dim(counts_liver)
# dim(col_data); dim(row_data)

MouseScialdoneLiver <- new("ExpressionSet",
                exprs = as.matrix(counts_liver),
                phenoData = new("AnnotatedDataFrame",
                                data = col_data),
                featureData = new("AnnotatedDataFrame",
                                  data = row_data),
                experimentData = new("MIAME",
                                     title = "Mouse Sciladone liver"))

save(MouseScialdoneLiver,
        file = "../data/all-rda/MouseScialdoneLiver.rda")
```


# Bulk RNA-seq

## Scialdone et al 2015, reads


```{r, eval = F}
library(data.table)
G1_data <- read.table("../data/MouseScialdoneESC/G1_htseq_count.txt");
G2M_data <- read.table("../data/MouseScialdoneESC/G2M_htseq_count.txt");
S_data <- read.table("../data/MouseScialdoneESC/S_htseq_count.txt");
colnames(G1_data) <- c("gene_names", "G1")
colnames(G2M_data) <- c("gene_names", "G2M")
colnames(S_data) <- c("gene_names", "S")

dim(G1_data)
dim(G2M_data)
dim(S_data)

counts_ESC <- merge(G1_data, G2M_data, by = "gene_names")
counts_ESC <- merge(counts_ESC, S_data, by = "gene_names")
head(counts_ESC)

row_data <- data.frame(gene_name = as.character(counts_ESC$gene_names),
                       stringsAsFactors = FALSE)
rownames(row_data) <- counts_ESC$gene_names

col_data <- data.frame(cycle_phase = colnames(counts_ESC)[-1])
rownames(col_data) <- col_data$cycle_phase

meta_list <- data.frame(labelDescription = c("cycle phase"),
                    row.names = colnames(col_data))

counts_ESC <- counts_ESC[,-1]

MouseScialdoneESC <- new("ExpressionSet",
                exprs = as.matrix(counts_ESC),
=======
## Blakeley 2015 Human embryo, counts

```{r eval=F}
data <- data.frame(fread("../data/GSE66507_human_blastocyst_rnaseq_counts.txt"));

gene_names <- data[,1];
reads <- data[,-1];
rownames(reads) <- gene_names;

row_data <- data.frame(gene_name = as.character(rownames(reads)),
                       stringsAsFactors = FALSE)
rownames(row_data) <- rownames(reads)


metadata1 <- colnames(reads);
metadata2 <- unlist(lapply(1:length(metadata1), function(l) return(strsplit(metadata1[l],"[.]")[[1]][1])))
metadata3 <- unlist(lapply(1:length(metadata2), function(l) return(substr(metadata2[l],1, nchar(metadata2[l])-1))))

col_data <- data.frame(
    cell_type = metadata3,
    cell_type_class = metadata2,
    cell_id = metadata1);

colnames(col_data) <- c("cell_type", "cell_type_class", "cell_id");
rownames(col_data) <- colnames(reads);

HumanBlakeleyEmbryo <- new("ExpressionSet",
                exprs = as.matrix(reads),
>>>>>>> 89b30f8a5b87ddb9b95a5af4ae098e115ea566be
                phenoData = new("AnnotatedDataFrame",
                                data = col_data),
                featureData = new("AnnotatedDataFrame",
                                  data = row_data),
                experimentData = new("MIAME",
<<<<<<< HEAD
                                     title = "Mouse Sciladone ESC"))

save(MouseScialdoneESC,
        file = "../data/all-rda/MouseScialdoneESC.rda")

```


=======
                                     title = "Blakeley 2015 Human Embryo"))

save(HumanBlakeleyEmbryo,
     file = "../data/all-rds/HumanBlakeleyEmbryo.rda")

```

>>>>>>> 89b30f8a5b87ddb9b95a5af4ae098e115ea566be


## Session information
```{r}
sessionInfo()
```

