---
title: "SMASH software application in time ordering- validation check"
author: "Kushal K Dey"
date: "January 14, 2016"
output: 
  html_document:
    toc: true
---

## Background

The goal of the study in this analysis is to apply the **SMASH** R package due to Tom Xing and Matthew Stephens for determining a sensible smoothing optimization  criterion for cell cycle data. The **SMASH** package fits nonparameteric models to 

$$ y_{i} = \mu_{i} + \epsilon_{i} $$

$$ \epsilon_{i} \sim N(0, \sigma^2_{i}) $$

Our goal then is to estimate the underlying parameter $\mu$. In the **cellcycleR** set up, we also have similar condition and smoother the $\mu$, the fit is expected to be better (the error variation will be smaller) as wavelet methods make use of the neighborhood behavior at any point. 

## Simulation from sinusoidal model

```{r, echo=TRUE, eval=TRUE}
library(cellcycleR)
library(wavethresh)
```


```{r echo=TRUE, eval=TRUE}

G <- 100;
num_cells <- 256;
amp_genes1 <- rep(10, G);
phi_genes1 <- rep(c(2,5), each=G/2);
sigma_genes1 <- rchisq(G, 4);
cell_times_sim <- sort(sample(seq(0,2*pi, 2*pi/(num_cells-1)), num_cells, replace=FALSE));
cycle_data <- t(sim_sinusoidal_cycle(G, amp_genes1, phi_genes1, sigma_genes1, cell_times_sim));

```

The sinusoidal expression pattern for $1$st gene 

```{r echo=TRUE, eval=TRUE}
plot(cycle_data[1,])
```

## Permutation of sample labels

Now we permute the labels of the times.

```{r echo=TRUE, eval=TRUE}
sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
cycle_data_reorder <- cycle_data[,sample_reorder];
```

We plot the expression pattern for $1$st gene.

```{r echo=TRUE, eval=TRUE}
plot(cycle_data_reorder[1,])
```

## Applying SMASH

We load the **SMASH** package.

```{r echo=TRUE, eval=TRUE}
library(smash)
mu.est.true <- suppressMessages(suppressWarnings(ashsmooth.gaus(cycle_data[1,])))
plot(mu.est.true, type = "l")

mu.est.reorder <- suppressMessages(suppressWarnings(ashsmooth.gaus(cycle_data_reorder[1,])))
plot(mu.est.reorder, type = "l")
```

We then find the deviation of true observations against the fitted mean value from the above method.

```{r echo=TRUE, eval=TRUE}
disp_reorder <- sum((cycle_data_reorder[1,] - mu.est.reorder)^2)
disp_true <- sum((cycle_data[1,] - mu.est.true)^2)
```

Check that the dispersion of actual observations from the fitted mean curve is smaller for the true order compared to the re-ordered time series data.

## Validation of procedure

But this was for one gene only. We can find the average dispersion across all the genes.

```{r echo=TRUE, eval=TRUE}
disp.true <- array(0,G);
smash_disp_true <- mean(sapply(1:G, function(g)
                               {
                                  mu.est.true <- suppressMessages(suppressWarnings(ashsmooth.gaus(cycle_data[g,])))
                                  out <- sum((cycle_data[g,] - mu.est.true)^2)
                                  return(out)
                               }))
```

Now we reorder the data and apply the same method $30$ times and plot the density of the dispersion values and check where this density function lies against the true value (corresponding to the true order) we obtained above.

```{r echo=TRUE, eval=TRUE}
library(parallel)
outdisp_set <- unlist(mclapply(1:20, function(l)
               {
                  sample_reorder <- sample(1:num_cells,num_cells, replace=FALSE);
                  cycle_data_reorder <- cycle_data[,sample_reorder];
                  disp <- mean(sapply(1:G, function(g)
                               {
                                  mu.est.reord <- suppressMessages(suppressWarnings(ashsmooth.gaus(cycle_data_reorder[g,])))
                                  out <- sum((cycle_data_reorder[g,] - mu.est.reord)^2)
                                  return(out)
                               }))
                  return(disp)
                }, mc.cores=detectCores()))

```

The dispersion set of values for re-ordered data (not true order) after repeating the re-ordering $20$ times 

```{r echo=TRUE, eval=TRUE}
print(outdisp_set)
```

The true value of dispersion corresponding to true ordering 

```{r echo=TRUE, eval=TRUE}
print(smash_disp_true)
```
