---
title: 'classtpx vs SVM on GTEx V6 data: (Thyroid + Adrenal)'
author: "Kushal K Dey"
date: "April 2, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Overview

In this script, we extract tissue samples from two tissues in the GTEx V6 samples and apply a classification due to Support Vector Machines (SVM) and the classtpx model. We would like to figure out which of the two methods has higher misclassification rate.

```{r warning=FALSE, message=FALSE}
library(data.table)
library(e1071)
library(classtpx)
library(maptpx)
```

## Data preparation (Thyroid + Adrenal Gland)

```{r message=FALSE, warning=FALSE}
gtex_data <- data.frame(fread("../data/GTEX_V6/cis_gene_expression.txt"));
matdata <- gtex_data[,-(1:2)];

tissue_labels=read.table("../data/GTEX_V6/samples_id.txt")[,3];

thyroid_labels <- which(as.character(tissue_labels) == 'Thyroid')
adrenal_labels <- which(as.character(tissue_labels) == 'Adrenal Gland');

pooled_data <- t(cbind(matdata[,thyroid_labels],matdata[,adrenal_labels]))
```

## maptpx application

We first perform `maptpx`.

```{r echo=TRUE, eval=FALSE}
Topic_clus <- maptpx::topics(pooled_data, K=2, tol=10);
save(Topic_clus, file="../rdas/gtex_thyroid_adrenal.rda");
```

We plot suing the Structure plot visualization.

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/gtex_thyroid_adrenal.rda"));

par(mfrow=c(1,1))
par(mar=c(4,2,2,2))
omega <- Topic_clus$omega;
annotation <- data.frame(
sample_id = paste0("X", c(1:NROW(omega))),
tissue_label = factor(tissue_labels[c(thyroid_labels, adrenal_labels)],
                      levels=c("Thyroid", "Adrenal Gland"))
)

rownames(omega) <- annotation$sample_id;
CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))

```

## Support Vector Machine (SVM) application 

We filter out the first $50$ samples from the Blood samples and the first $50$ samples from the Pancreas tissues and then perform classification of the other samples based on these two sets of samples. 

### On counts

```{r echo=TRUE, eval=TRUE}

gene_names <- read.table("../data/GTEX_V6/gene_names_GTEX_V6.txt");
gene_names <- substring(gene_names[,1],1,15);

training_data <- t(cbind(matdata[,thyroid_labels[1:50]],matdata[,adrenal_labels[1:50]]))

test_data <- t(cbind(matdata[,thyroid_labels[-(1:50)]],matdata[,adrenal_labels[-(1:50)]]))

tissue_type <- c(rep("Thyroid",50), rep("Adrenal Gland",50));

training.data.frame <- cbind.data.frame(training_data, tissue_type);
colnames(training.data.frame) <- c(gene_names, "tissue_type");

model <- svm(tissue_type ~ ., data = training.data.frame)

test_class_svm <- predict(model, test_data)

tab_class_svm <- table(test_class_svm, c(rep("Thyroid", length(thyroid_labels)-50), rep("Adrenal Gland", length(adrenal_labels)-50)))

print(tab_class_svm)

misclass_svm <- (tab_class_svm[1,2]+tab_class_svm[2,1])/sum(tab_class_svm)
misclass_svm
```

### On voom transformed counts

```{r echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
voom_matdata <- data.frame(fread("../data/GTEX_V6/voom_cis_gene_expression.txt"))
voom_matdata <- voom_matdata[,-1]

training_data <- t(cbind(voom_matdata[,thyroid_labels[1:50]],voom_matdata[,adrenal_labels[1:50]]))

test_data <- t(cbind(voom_matdata[,thyroid_labels[-(1:50)]],voom_matdata[,adrenal_labels[-(1:50)]]))

tissue_type <- c(rep("Thyroid",50), rep("Adrenal Gland",50));

training.data.frame <- cbind.data.frame(training_data, tissue_type);
colnames(training.data.frame) <- c(gene_names, "tissue_type");

model <- svm(tissue_type ~ ., data = training.data.frame)

test_class_svm <- predict(model, test_data)

tab_class_svm <- table(test_class_svm, c(rep("Thyroid", length(thyroid_labels)-50), rep("Adrenal Gland", length(adrenal_labels)-50)))

print(tab_class_svm)

misclass_svm <- (tab_class_svm[1,2]+tab_class_svm[2,1])/sum(tab_class_svm)
misclass_svm
```


## classtpx (theta.fix) application

```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,50), rep(2,50));
known_samples <- c(rep(1,50), length(thyroid_labels) + (1:50));

Topic_clus <- classtpx::class_topics(
    pooled_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    shrink=FALSE,
    tol=0.01)

save(Topic_clus, file="../rdas/thyroid_adrenal_theta_fix_classtpx.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/thyroid_adrenal_theta_fix_classtpx.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Thyroid", "Adrenal Gland")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(rep("Thyroid", length(thyroid_labels)), 
                              rep("Adrenal Gland", length(adrenal_labels)))[c(51:length(thyroid_labels), (length(thyroid_labels)+51):dim(omega)[1])],
                            test_class_classtpx[c(51:length(thyroid_labels), (length(thyroid_labels)+51):dim(omega)[1])]);

print(tab_class_classtpx)

misclass_classtpx <- (tab_class_classtpx[1,2]+tab_class_classtpx[2,1])/sum(tab_class_classtpx)
misclass_classtpx
```

## classtpx (omega.fix) application

## classtpx (theta.fix) application

```{r echo=TRUE, eval=FALSE}
class_labs <- c(rep(1,50), rep(2,50));
known_samples <- c(rep(1,50), length(thyroid_labels) + (1:50));

Topic_clus <- classtpx::class_topics(
    pooled_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="omega.fix",
    shrink=FALSE,
    tol=0.01)

save(Topic_clus, file="../rdas/thyroid_adrenal_omega_fix_classtpx.rda")
```

```{r echo=TRUE, eval=TRUE}
Topic_clus <- get(load("../rdas/thyroid_adrenal_omega_fix_classtpx.rda"));

omega <- Topic_clus$omega;

CountClust::StructureGGplot(omega = as.matrix(omega),
                    annotation = annotation,
                    palette = RColorBrewer::brewer.pal(8, "Accent"),
                    yaxis_label = "Tissue type",
                    order_sample = TRUE,
                    axis_tick = list(axis_ticks_length = .1,
                                     axis_ticks_lwd_y = .1,
                                     axis_ticks_lwd_x = .1,
                                     axis_label_size = 7,
                                     axis_label_face = "bold"))


test_class_classtpx <- c("Thyroid", "Adrenal Gland")[apply(omega,1,function(x) which.max(x))];
tab_class_classtpx <- table(c(rep("Thyroid", length(thyroid_labels)), 
                              rep("Adrenal Gland", length(adrenal_labels)))[c(51:length(thyroid_labels), (length(thyroid_labels)+51):dim(omega)[1])],
                            test_class_classtpx[c(51:length(thyroid_labels), (length(thyroid_labels)+51):dim(omega)[1])]);

print(tab_class_classtpx)

misclass_classtpx <- (tab_class_classtpx[1,2]+tab_class_classtpx[2,1])/sum(tab_class_classtpx)
misclass_classtpx
```
