---
title: "classtpx on Treutlin et al 2014"
author: "Kushal K Dey"
date: "March 22, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Introduction

We apply the `classtpx` package for topic model classfication on the [Treutlin et al 2014](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4145853/pdf/nihms568867.pdf). The data for this package is saved as an ExpressionSet object at the following [link](https://github.com/jhsiao999/singleCellRNASeqMouseTreutleinLung). We first load the data.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
# devtools::install_github("jhsiao999/singleCellRNASeqMouseTreutleinLung", force=TRUE)
library("singleCellRNASeqMouseTreutlein")
data("MouseTreutleinLung")
```

```{r}
counts_data <- t(Biobase::exprs(MouseTreutleinLung));
pheno_metadata <- pData(MouseTreutleinLung);
table(pheno_metadata[,1])
```

## maptpx application

```{r echo=TRUE, eval=FALSE}
Topic_clus <- maptpx::topics(counts_data, 3, tol=0.1);
save(Topic_clus, file="../rdas/treutlin_topic_fit_3_maptpx.rda")
```

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.show="asis", dpi=144, fig.width=3, fig.height=5}
Topic_clus <- get(load(file="../rdas/treutlin_topic_fit_3_maptpx.rda"))

omega <- Topic_clus$omega;

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(paste0(pheno_metadata$cell_type, "_", pheno_metadata$replicate),
                        levels=rev(c("E14.5_1", "E16.5_1", "E18.5_1","E18.5_2","E18.5_3","adult_replicate"))
))

rownames(omega) <- annotation$sample_id;


CountClust::StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Cell cycle phase",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))
```

It seems the ideal fit would be $2$ clusters. We now consider one class to be the adult replicate and the other would be E14.5.

## classtpx model

## omega.fix method 


```{r}
known_samples <- c(which(pheno_metadata$cell_type=="E14.5"), which(pheno_metadata$cell_type=="adult"));
class_labs <- c(rep(1, length(which(pheno_metadata$cell_type=="E14.5"))), rep(2,length(which(pheno_metadata$cell_type=="adult"))));
```

### K=2 method

```{r echo=TRUE, eval=FALSE}
Topic_clus <- classtpx::class_topics(
    counts_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="omega.fix",
    tol=0.01,
    shrink=FALSE)

save(Topic_clus, file="../rdas/treutlin_topic_fit_2_classtpx_omega_fix.rda")
```

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.show="asis", dpi=144, fig.width=3, fig.height=5}
Topic_clus <- get(load(file="../rdas/treutlin_topic_fit_2_classtpx_omega_fix.rda"))

omega <- Topic_clus$omega;

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(paste0(pheno_metadata$cell_type, "_", pheno_metadata$replicate),
                        levels=rev(c("E14.5_1", "E16.5_1", "E18.5_1","E18.5_2","E18.5_3","adult_replicate"))
))

rownames(omega) <- annotation$sample_id;


CountClust::StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Cell cycle phase",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))
```


### K=3 method

```{r echo=TRUE, eval=FALSE}
Topic_clus <- classtpx::class_topics(
    counts_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="omega.fix",
    tol=0.01,
    shrink=FALSE)

save(Topic_clus, file="../rdas/treutlin_topic_fit_2_classtpx_omega_fix.rda")
```

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.show="asis", dpi=144, fig.width=3, fig.height=5}
Topic_clus <- get(load(file="../rdas/treutlin_topic_fit_2_classtpx_omega_fix.rda"))

omega <- Topic_clus$omega;

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(paste0(pheno_metadata$cell_type, "_", pheno_metadata$replicate),
                        levels=rev(c("E14.5_1", "E16.5_1", "E18.5_1","E18.5_2","E18.5_3","adult_replicate"))
))

rownames(omega) <- annotation$sample_id;


CountClust::StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Cell cycle phase",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))
```

## theta.fix method

### K=2

```{r echo=TRUE, eval=FALSE}
Topic_clus <- classtpx::class_topics(
    counts_data, 
    K=2, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    tol=0.01,
    shrink=FALSE)

save(Topic_clus, file="../rdas/treutlin_topic_fit_2_classtpx_theta_fix.rda")
```

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.show="asis", dpi=144, fig.width=3, fig.height=5}
Topic_clus <- get(load(file="../rdas/treutlin_topic_fit_2_classtpx_theta_fix.rda"))

omega <- Topic_clus$omega;

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(paste0(pheno_metadata$cell_type, "_", pheno_metadata$replicate),
                        levels=rev(c("E14.5_1", "E16.5_1", "E18.5_1","E18.5_2","E18.5_3","adult_replicate"))
))

rownames(omega) <- annotation$sample_id;


CountClust::StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Cell cycle phase",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))
```

### K=3

```{r echo=TRUE, eval=FALSE}
Topic_clus <- classtpx::class_topics(
    counts_data, 
    K=3, 
    known_samples = known_samples,
    class_labs = class_labs,
    method="theta.fix",
    tol=0.01,
    shrink=FALSE)

save(Topic_clus, file="../rdas/treutlin_topic_fit_3_classtpx_theta_fix.rda")
```

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.show="asis", dpi=144, fig.width=3, fig.height=5}
Topic_clus <- get(load(file="../rdas/treutlin_topic_fit_3_classtpx_theta_fix.rda"))

omega <- Topic_clus$omega;

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(paste0(pheno_metadata$cell_type, "_", pheno_metadata$replicate),
                        levels=rev(c("E14.5_1", "E16.5_1", "E18.5_1","E18.5_2","E18.5_3","adult_replicate"))
))

rownames(omega) <- annotation$sample_id;


CountClust::StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Cell cycle phase",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))
```

## Conclusion

the classtpx model for K=2 clearly shows a continuous gradient from E14.5 stage to adult replicate stage. When we perform K=3, the continuity in gradient is preserved.

