---
title: 'Yoav cellcycle gene annotations: full data'
author: "Kushal K Dey"
date: "November 11, 2015"
output: html_document
---


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(qtlcharts)
library(cellcycleR)
library(data.table)
library(limma)
library(mygene)
library(knitr)
```

## Objective 

In this script, we annotate the genes that were found to be sinusoidal under the cellcycleR method, after applying it on Yoav's data. It must be noted that so far, we have only focussed on the cell cycle gene list provided in Macosko paper (see [here]) for mouse retinal cells. It may very well happen that some of these genes do not shpw cell cycle patterns for iPSCs and it can also happen that there are other genes outside this list that show cell cycle patterns. The latter can lead to discovery of new cell cycle genes for the iPSCs or stem cells in general.

Despite this loophole, we perform the cellcycleR method on the listed cell cycle genes and then apply the gene annotations on this reduced list. We ran cellcycleR method on Yoav's data, with the details of the results in our [webpage](http://jhsiao999.github.io/singleCell-method/project/analysis/yoav_cellcycleR.html).

Load the data

```{r echo=TRUE, eval=TRUE}
setwd('/Users/kushal/Documents/singleCell-method/project/analysis')
colnames <- colnames(read.table("../data/batch_removed_counts_all_genes.txt"));
molecules_single <- data.frame(fread("../data/batch_removed_counts_all_genes.txt"), row.names =1);
colnames(molecules_single) <- colnames
cycle_counts_data <- molecules_single;
cycle_voom_data <- voom(cycle_counts_data)$E;
cycle_data_norm <- apply(cycle_voom_data,2,function(x)  return (x-mean(x))/sd(x))
cycle_data_norm <- cycle_data_norm[, which(colSums(cycle_data_norm)!=0)]

out <- get(load(file="../rdas/cell_order_ipsc_full.rda"));
amp_genes <- out$amp;
sd_genes <- out$sigma;
phi_genes <- out$phi;

ESS <- amp_genes^2; RSS <- sd_genes^2

SNR <- ESS/RSS;

snr_high_indices <- which(SNR > 3);

cycle_data_norm_sinusoidal <- cycle_data_norm[,snr_high_indices];

dim(cycle_data_norm_sinusoidal)

rm(SNR); rm(amp_genes); rm(sd_genes); rm(phi_genes);

```

Load the cellcycleR output

```{r echo=TRUE, eval=TRUE}
 out2 <- get(load(file="../rdas/cell_order_ipsc_sinusoidal_full.rda"));
 
 amp_genes <- out2$amp;
 sd_genes <- out2$sigma;
 phi_genes <- out2$phi;
 
 ESS <- amp_genes^2; RSS <- sd_genes^2
 
 SNR <- ESS/RSS;

```

## Annotations

Extracting the top 100 gene names with highest SNR.

```{r echo=TRUE, eval=TRUE}
sinusoidal_gene_names <- colnames(cycle_data_norm_sinusoidal);
imp_genes <- sinusoidal_gene_names[order(SNR, decreasing=TRUE)]
top_genes_100 <- imp_genes[1:200];
```

We now annotate these top $100$ genes using **mygene** package.

```{r echo=TRUE, eval=TRUE}
out <- queryMany(top_genes_100,  scopes="ensembl.gene", fields=c("name","summary"), species="human");

kable(as.data.frame(out))

```
