---
title: 'Yoav cellcycleR post processing: annotating genes'
author: "Kushal K Dey"
date: "November 5, 2015"
output: 
  html_document:
    toc: true
---

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(qtlcharts)
library(cellcycleR)
library(data.table)
library(limma)
library(mygene)
library(knitr)
```

## Objective 

In this script, we annotate the genes that were found to be sinusoidal under the cellcycleR method, after applying it on Yoav's data. It must be noted that so far, we have only focussed on the cell cycle gene list provided in Macosko paper (see [here]) for mouse retinal cells. It may very well happen that some of these genes do not shpw cell cycle patterns for iPSCs and it can also happen that there are other genes outside this list that show cell cycle patterns. The latter can lead to discovery of new cell cycle genes for the iPSCs or stem cells in general.

Despite this loophole, we perform the cellcycleR method on the listed cell cycle genes and then apply the gene annotations on this reduced list. We ran cellcycleR method on Yoav's data, with the details of the results in our [webpage](http://jhsiao999.github.io/singleCell-method/project/analysis/yoav_cellcycleR.html).

Load the data

```{r echo=TRUE, eval=TRUE}
setwd('/Users/kushal/Documents/singleCell-method/project/analysis')
molecules_single_cell_cycle <- read.table("../data/molecules_ipsc_single_cell_cycle.txt");
cycle_counts_data <- t(molecules_single_cell_cycle);
cycle_voom_data <- voom(cycle_counts_data)$E;
cycle_data_norm <- apply(cycle_voom_data,2,function(x)  return (x-mean(x))/sd(x))
cycle_data_norm <- cycle_data_norm[, -which(colSums(cycle_data_norm)==0)]

out <- get(load(file="../rdas/cell_order_ipsc.rda"));
amp_genes <- out$amp;
sd_genes <- out$sigma;
phi_genes <- out$phi;

ESS <- amp_genes^2; RSS <- sd_genes^2

SNR <- ESS/RSS;

snr_high_indices <- which(SNR > 1);

cycle_data_norm_sinusoidal <- cycle_data_norm[,snr_high_indices];

dim(cycle_data_norm_sinusoidal)

rm(SNR); rm(amp_genes); rm(sd_genes); rm(phi_genes);

```


Load the cellcycleR output

```{r echo=TRUE, eval=TRUE}
out2 <- get(load(file="../rdas/cell_order_ipsc_sinusoidal.rda"));

amp_genes <- out2$amp;
sd_genes <- out2$sigma;
phi_genes <- out2$phi;

ESS <- amp_genes^2; RSS <- sd_genes^2

SNR <- ESS/RSS;

```

## Annotations

Extracting the top 100 gene names with highest SNR.

```{r echo=TRUE, eval=TRUE}
sinusoidal_gene_names <- colnames(cycle_data_norm_sinusoidal);
imp_genes <- sinusoidal_gene_names[order(SNR, decreasing=TRUE)]
top_genes_100 <- imp_genes[1:100];
```

We now annotate these top $100$ genes using **mygene** package.

```{r echo=TRUE, eval=TRUE}
out <- queryMany(top_genes_100,  scopes="ensembl.gene", fields=c("name","summary"), species="human");

kable(as.data.frame(out))

```

## Observations 

It seems to me that cellcycleR is picking up some really important genes with known cell cycle functions. There are a quite a few CDC protein related genes and a few cyclin related ones, which are most important driving factors of cell cycle. Also, going by the names of the genes detected, some of them clearly are said to be related to cell cycle and cell division. If we compare to the gene annotations from the admixture method which also separated out the G2.M and M phases from the rest, we did not annotate so many relevant genes. This seems to be a very positive finding.

