---
title: 'Fash for variable selection: Example 1'
author: "Kushal K Dey"
date: "March 16, 2016"
output: html_document
---

## Overview

In this script, we experiment with the `fash` software dur to Mengyin Lu and 
apply it on a simulated RNA-seq counts data to observe how well it performs in variable weighting. We consider a `classtpx` framework where we have some 
samples coming fom known groups and we want to figure out which genes are 
informative in separating out the samples between the two classes. Note that 
these are the likely genes that will drive the clusters in the GoM (grade of 
membership) model fit in `classtpx`. So this could serve as an important pre-
processing tool.

```{r echo=TRUE, eval=TRUE}
n.out <- 200
omega_sim <- rbind( cbind( rep(1, n.out), rep(0, n.out)), 
                    cbind( rep(0, n.out), rep(1, n.out)),
                    cbind( seq(0.6, 0.4, length.out = n.out), 
                           1- seq(0.6, 0.4,length.out=n.out)) )
dim(omega_sim)

freq <- rbind( c(0.1, 0.2, rep(0.70/98, 98)),
               c(rep(0.70/98,49), 0.1, rep(0.70/98,49), 0.2) )
str(freq)

counts <- t( do.call(cbind,
                     lapply(1:dim(omega_sim)[1], 
                            function(x) 
                                rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)

```

Extracting the data already known to be in classes.

```{r echo=TRUE, eval=TRUE}
counts_class <- counts[1:(2*n.out),];
class_labs <- c(rep(1,n.out), rep(2, n.out));
```

We compute the F-statistic scores for all the genes.

```{r echo=TRUE, eval=TRUE}
feature_var_within_class <- parallel::mclapply(1:dim(counts_class)[2], 
            function(l) {
                    ll <- tapply(counts_class[,l], class_labs, var);
                    num_id <-sum(as.numeric(table(class_labs)-1)*ll)/(length(class_labs)-length(unique(class_labs)));
                    })

feature_var_between_class <- parallel::mclapply(1:dim(counts_class)[2], 
                          function(l) {
                              ll <- tapply(counts_class[,l], class_labs, mean);
                              return(var(ll))
                            })
```

```{r echo=TRUE, eval=TRUE}
f.ratio <- unlist(lapply(1:dim(counts_class)[2], function(l) return (feature_var_between_class[[l]]/feature_var_within_class[[l]])));

out <- fash((f.ratio+1e-20), (length(unique(class_labs))-1), (length(class_labs)-length(unique(class_labs))), g=NULL, optmethod="mixEM", prior="nullbiased", oneside=FALSE)

out <- fash((f.ratio+1e-20), (length(unique(class_labs))-1), (length(class_labs)-length(unique(class_labs))))

exp(out$PosteriorMean.logf)

out$PosteriorMean.f


```

