---
title: 'Assume correlated gene expression'
author: "Joyce Hsiao"
date: "January 26, 2016"
output: 
  html_document:
    toc: true
---

## Background and objective

It is reasonable to assume that gene expression patterns are correlated among genes that regulate similar pathways. When modeling cell-cycle, genes that are involved in regulating the same cell-phase are likely to be correlated in their expression patterns. Here we consider a simple example where mean gene expression levels follow a sinusoidal pattern and a simple non-sinusoidal pattern. Gene expression levels are assumed to be correlated and follow a homogeneous covarying pattern across all possible gene pairs.

Note that assumption of homogeneous correlation between genes is employed for a matter of convenience. 


## Load packages

Load packages.

```{r echo=FALSE, eval=TRUE}
library(cellcycleR)
library(wavethresh)
```


## Case 1: Sinusoidal model, correlated expression

Set parameters.

```{r}
G <- 100
num_cells <- 256
amp_genes <- rep(10, G)
phi_genes <- rep(c(2,5), each=G/2)
sigma_genes <- rchisq(G, 4)
cell_times_sim <- seq(0,2*pi, length.out=num_cells)
cycle_data <- sim_sinusoidal_cycle(G, amp_genes, phi_genes,
                                      sigma_genes, cell_times_sim,
                                      correlation = .3)
```


Randomly select some genes and plot out the simulated expression pattern.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data[ ,which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Simulated pattern", outer = TRUE, line = -1)
```

Shuffle labels of cell-phase (pi) between cells.

```{r echo=TRUE, eval=TRUE}
sample_reorder <- sample(1:num_cells, num_cells, replace=FALSE)
cell_times_reorder <- cell_times_sim[sample_reorder]
cycle_data_reorder <- cycle_data[sample_reorder,]

par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ ,which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
```


### Sinusoidal cellcycleR 

```{r echo=FALSE, eval=TRUE}
print(system.time(sinu_trend_fit_sinu <- sin_cell_ordering_class(cycle_data_reorder, 
                                           celltime_levels = 100, num_iter = 100) ) 
  )
```

Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(sinu_trend_fit_sinu$cell_times), 
                           which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Sinusoidal model", outer = TRUE, line = -1)
```


### Smoothing splines fitting

```{r echo=FALSE, eval=TRUE}
print( system.time(
      splines_fit <- np_cell_ordering_class(cycle_data_reorder, 
                                            celltime_levels = 256, 
                                            num_iter = 500, 
                                            method = "B-spline")
    ) )
str(splines_fit)
```

Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(splines_fit$cell_times), which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Smoothing splines", outer = TRUE, line = -1)
```



### Wavelet

```{r echo=FALSE, eval=TRUE}
print( system.time(
      wavelet_fit <- np_cell_ordering_class(cycle_data_reorder, 
                                            celltime_levels = 256, 
                                            num_iter = 500, 
                                            method = "Wavelet")
    ) )
str(wavelet_fit)
```

Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(wavelet_fit$cell_times), which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Wavelet", outer = TRUE, line = -1)
```


## Case 2: Non-sinusoidal model, correlated expression

Set parameters.

```{r}
G <- 100
num_cells <- 256
phi_genes <- seq(1, pi, length.out=G)
cell_times_sim <- seq(0,2*pi, length.out=num_cells)
correlation <- .3
Sigma <- diag(G)
Sigma[upper.tri(Sigma)] <- correlation^2
Sigma[lower.tri(Sigma)] <- correlation^2

cycle_data <- matrix(0,num_cells,G)
for(s in 1:num_cells){
  	 cycle_data[s,] <- 10*cell_times_sim[s] + 30*sin(2*cell_times_sim[s] + phi_genes) + 
  	               mvrnorm(n = 1, mu = rep(0, G), Sigma = Sigma)
}
```

Shuffle labels of cell-phase (pi) between cells.

```{r echo=TRUE, eval=TRUE}
sample_reorder <- sample(1:num_cells, num_cells, replace=FALSE)
cell_times_reorder <- cell_times_sim[sample_reorder]
cycle_data_reorder <- cycle_data[sample_reorder,]

par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ ,which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
```


### Sinusoidal cellcycleR 

```{r echo=TRUE, eval=TRUE}
system.time(
  sinusoidal_fit <- sin_cell_ordering_class(cycle_data_reorder, 
                                            celltime_levels = 256, num_iter = 300) 
  )
```


Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(sinusoidal_fit$cell_times), which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Sinusoidal model", outer = TRUE, line = -1)
```


### Smoothing splines fitting

```{r echo=TRUE, eval=TRUE}
system.time(
  splines_fit <- np_cell_ordering_class(cycle_data_reorder, 
                                        celltime_levels = 256, 
                                        num_iter = 300, 
                                        method = "B-spline")
  )
```


Recovered cell-order.

```{r echo=TRUE, eval=TRUE}
par(mfrow = c(2,2))
for (which_gene in c(1, 30, 60)) {
  plot(cycle_data_reorder[ order(splines_fit$cell_times), which_gene], type="l", 
       main = paste("Gene no.", which_gene),
       xlab = "Cell ordered by cell-phase",
       ylab = "Gene expression")
}
title("Smoothing splines", outer = TRUE, line = -1)
```





