---
title: 'Human iPSC: PCA vs. sinusoidal modeling'
author: "Joyce Hsiao"
date: "2016-01-28"
output: 
  html_document:
    toc: true
---


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(qtlcharts)
library(CountClust)
library(parallel)
library(cellcycleR)
library(data.table)
library(binhf)
library(vioplot)
library(limma)
library(readxl)
library(Humanzee)
library(mygene)
library(knitr)
```


## Background and objectives

After fitting sinusoidal model to [Gilad2015][singleCellSeq] iPSC NA19239 ([link1][link1], we fitted PCA to the same data and compare fitting results.

[link1]: http://jhsiao999.github.io/singleCell-method/project/analysis/yoav_cellcycleR_cellcycle_genes-2016-01-28.html

[singleCellSeq]: http://jdblischak.github.io/singleCellSeq/analysis/


## Conclusions


We fit the model under the scenarios of 1) all annotated cell-cycle genes, 2) high SNR from cell-cycle genes, 3) best set from the high-SNR genes. In all scenarios, PCA performs poorly... 


## Data prepration


### Import final batch-corrected iPSC data

```{r echo=TRUE, eval=TRUE}
molecules_final <- read.table("../data/gilad-2015/molecules-final.txt",
                              header = TRUE,
                              stringsAsFactors = FALSE)
anno_filter <- read.table("../data/gilad-2015/annotation-filter.txt",
                          header = TRUE,
                          stringsAsFactors = FALSE)
dim(molecules_final)
table(anno_filter$individual)
```

Extract only data of one individual

```{r}
molecules_final_subset <- molecules_final[ , anno_filter$individual == "NA19239"]
dim(molecules_final_subset)

molecules_final_subset <- molecules_final_subset
```


Import cell-cycle genes

```{r}
cellcycle_genes <- read.table("../data/gilad-2015/cellcyclegenes.txt",
                               header = TRUE,
                               sep = "\t",
                               stringsAsFactors = FALSE)
colnames(cellcycle_genes) <- c("G1.S","S","G2","G2.M","M.G1")
```


Extract only cell-cycle genes

```{r}
which_cell_cycle <- which (rownames(molecules_final_subset) %in% unlist(cellcycle_genes))
cycle_data <- t(molecules_final_subset[which_cell_cycle, ])
dim(cycle_data)
```

Standardize expression levels into z-scores for for gene. This step is for easy visualization of expression levels in plots.

```{r}
cycle_data_normed <- apply(cycle_data, 2, 
                            function(x)  return (x-mean(x))/sd(x))
```



## Fitting on cell-cycle genes

```{r}
if (file.exists("../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28.rda")) {
    load("../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28.rda")
} else {
results <- sin_cell_ordering_class(cycle_data_normed, 
                                   celltime_levels = 300,
                                   num_iter = 300)
save(results, 
     file ="../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28.rda")
}
str(results)
```


Post-processing of cell-phase order.

```{r}
cell_order_full <- cell_ordering_full(results$signal_intensity, dim(molecules_final)[2])
str(cell_order_full)
```


### Model estimates

We look at the plots of the amplitudes, phases and the non signal variation of the genes.

```{r}
amp_genes <- results$amp
sd_genes <- results$sigma
phi_genes <- results$phi

par(mfrow=c(2,2))
plot(density(phi_genes), col="red", 
     main="Density plot of the phases")
plot(density(amp_genes[-which.max(amp_genes)]), col="red", 
     main="Density plot of the amplitudes")
plot(density(sd_genes[-which.max(sd_genes)]), col="red", 
     main="Density plot of the non-signal sd")

ESS <- amp_genes^2
RSS <- sd_genes^2
SNR <- ESS/RSS
plot(SNR, col="red", pch=20, lwd=1)
```


### PCA vs. Re-ordered expression pattern

After re-ordering.

```{r}
new_cell_order <- binhf::shift(order(cell_order_full), 130, dir = "right")
iplotCurves(t(cycle_data_normed[new_cell_order, ]) )

pca_cycle <- prcomp(cycle_data_normed[new_cell_order, ], center=TRUE, scale. = TRUE)
par(mfrow= c(2,2))
plot(pca_cycle$x[,1], pca_cycle$x[,2], pch=20, lwd=0.01)
plot(pca_cycle$x[,1], pca_cycle$x[,3], pch=20, lwd=0.01)
plot(pca_cycle$x[,2], pca_cycle$x[,3], pch=20, lwd=0.01)
```



## Re-fitting on high SNR genes

We extract the high SNR genes as they seem to have sinusoidal patterns and repeat the procedure again. 

```{r}
snr_high_indices <- which(SNR > .1)
cycle_data_normed_high_snr <- cycle_data_normed[ ,snr_high_indices]
dim(cycle_data_normed_high_snr)
```

Modeing fitting...

```{r}
if (file.exists("../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28-high-snr.rda")) {
    load("../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28-high-snr.rda")
} else {
results_high_snr <- sin_cell_ordering_class(cycle_data_normed_high_snr, 
                                        celltime_levels = 300, num_iter=200)
save(results_high_snr, 
     file = "../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28-high-snr.rda")
}
```

Post-processing

```{r}
cell_order_full <- cell_ordering_full(results_high_snr$signal_intensity,
                                      dim(cycle_data_normed_high_snr)[2])
```


### Model estimates

We plot the same features as above and check for the robustness. We needed to shift the cell order so as to compare with previous plot on all genes as the method is non-identifiable upto a rotation. 

```{r echo=TRUE, eval=TRUE}
amp_genes <- results_high_snr$amp;
sd_genes <- results_high_snr$sigma;
phi_genes <- results_high_snr$phi;

par(mfrow = c(2,2))
plot(density(phi_genes), col="red", main="Density plot of the phases")
plot(density(amp_genes[-which.max(amp_genes)]), col="red", main="Density plot of the amplitudes")
plot(density(sd_genes[-which.max(sd_genes)]), col="red", main="Density plot of the non-signal sd")

ESS <- amp_genes^2; RSS <- sd_genes^2
SNR <- ESS/RSS;
plot(SNR, col="red", pch=20, lwd=1)
```

### PCA vs. Re-ordered expression patterns

```{r}
new_cell_order <- binhf::shift(order(cell_order_full), 130, dir = "right")
iplotCurves(t(cycle_data_normed_high_snr[new_cell_order, ]) ) 
pca_cycle <- prcomp(cycle_data_normed_high_snr[new_cell_order, ], 
                    center=TRUE, scale. = TRUE);
par(mfrow= c(2,2))
plot(pca_cycle$x[,1], pca_cycle$x[,2], pch=20, lwd=0.01)
plot(pca_cycle$x[,1], pca_cycle$x[,3], pch=20, lwd=0.01)
plot(pca_cycle$x[,2], pca_cycle$x[,3], pch=20, lwd=0.01)
```


## Re-fitting one more time...

```{r}
snr_high_indices <- which(SNR > .5)
cycle_data_normed_high_snr_high <- cycle_data_normed_high_snr[ ,snr_high_indices]
dim(cycle_data_normed_high_snr_high)
```

Modeing fitting...

```{r}
if (file.exists("../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28-high-snr-high.rda")) {
    load("../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28-high-snr-high.rda")
} else {
results_high_snr_high <- sin_cell_ordering_class(cycle_data_normed_high_snr_high, 
                                        celltime_levels = 300, num_iter=200)
save(results_high_snr_high, 
     file = "../rdas/yoav_cycleR_cellcycle_genes-2016-01-28/cellorder-ipsc-2016-01-28-high-snr0high.rda")
}
```

Post-processing

```{r}
cell_order_full <- cell_ordering_full(results_high_snr_high$signal_intensity,
                                      dim(cycle_data_normed_high_snr)[2])
```


### Model estimates

We plot the same features as above and check for the robustness. We needed to shift the cell order so as to compare with previous plot on all genes as the method is non-identifiable upto a rotation. 

```{r echo=TRUE, eval=TRUE}
amp_genes <- results_high_snr_high$amp;
sd_genes <- results_high_snr_high$sigma;
phi_genes <- results_high_snr_high$phi;

par(mfrow = c(2,2))
plot(density(phi_genes), col="red", main="Density plot of the phases")
plot(density(amp_genes[-which.max(amp_genes)]), col="red", main="Density plot of the amplitudes")
plot(density(sd_genes[-which.max(sd_genes)]), col="red", main="Density plot of the non-signal sd")

ESS <- amp_genes^2; RSS <- sd_genes^2
SNR <- ESS/RSS;
plot(SNR, col="red", pch=20, lwd=1)
```

### PCA vs. Re-ordered expression patterns

```{r}
new_cell_order <- binhf::shift(order(cell_order_full), 130, dir = "right")
iplotCurves(t(cycle_data_normed_high_snr_high[new_cell_order, ]) ) 
pca_cycle <- prcomp(cycle_data_normed_high_snr_high[new_cell_order, ], 
                    center=TRUE, scale. = TRUE);
par(mfrow= c(2,2))
plot(pca_cycle$x[,1], pca_cycle$x[,2], pch=20, lwd=0.01)
plot(pca_cycle$x[,1], pca_cycle$x[,3], pch=20, lwd=0.01)
plot(pca_cycle$x[,2], pca_cycle$x[,3], pch=20, lwd=0.01)
```



## Session information

```{r}
sessionInfo()
```

